---
title: "Gene-wise Bayesian DE with brmDE"
author: "Stefano Mangiola"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene-wise Bayesian DE with brmDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Fitting chunks run only when CmdStan is available (same gate as sccomp).
run_vignette <- instantiate::stan_cmdstan_exists()
```

# Overview

`brmDE` fits one gene at a time with [brms](https://paul-buerkner.github.io/brms/), then composes those fits into an [HPCell](https://github.com/MangiolaLaboratory/HPCell) `targets` pipeline. The typical flow is:

1. scale with TMM (`tidybulk::scale_abundance()`) and set `offset = log(1 / multiplier)`
2. estimate edgeR **dispersion** once on the full matrix (`estimate_dispersion()`)
3. fit each gene with `estimate_gene()`, using the dispersion to inform the negative binomial shape prior
4. optionally run posterior **hypothesis** tests and **adjust** for nuisance covariates

`estimate_gene()` never computes the offset or the dispersion itself. Those are inputs.

Fitting chunks below are gated by `instantiate::stan_cmdstan_exists()`, so the vignette still builds when CmdStan is missing.

# Install CmdStan

`brmDE` uses `cmdstanr` (not on CRAN) for the default Stan backend. Three steps:

1. R package install of `brmDE`
2. R install of `cmdstanr`
3. `cmdstanr` install of CmdStan itself

```{r install, eval = FALSE}
# from this repository
devtools::install()

install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev/", getOption("repos"))
)

cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

```{r packages, message = FALSE}
library(brmDE)
library(tidybulk)
library(tidySummarizedExperiment)
library(SummarizedExperiment)
data("airway", package = "airway")

airway$dex <- relevel(factor(airway$dex), ref = "untrt")
airway$cell <- droplevels(airway$cell)
```

# One gene, with dispersion

For a single gene you still prepare the offset and dispersion on a multi-gene matrix. After those columns exist, `filter()` the feature set you want to fit. These steps do not need CmdStan.

Get the TMM `multiplier` from `tidybulk::scale_abundance()`, then form the library-size offset as `log(1 / multiplier)` with `mutate()` from tidySummarizedExperiment (same convention as the immuneBodyMap / HPCell pipeline).

```{r one-gene-dispersion, message = FALSE}

se <- airway |>
  identify_abundant(formula_design = ~ dex) |>
  scale_abundance(method = "TMM") |>
  # library size factor is the reciprocal of the multiplier
  mutate(offset = log(1 / multiplier))

se <- se |> estimate_dispersion( formula = ~ dex + (1 | cell))

# One rowData column per estimate, both named by arguments of the same name:
#   dispersion                 = tagwise phi_g
#   dispersion_degrees_freedom = d_eff = df.residual + prior.df
head(as.data.frame(rowData(se)[, c("dispersion", "dispersion_degrees_freedom")]))
head(se$offset)

# Choose genes only after offset and dispersion are in the SE metadata
se_gene <- se |> filter(.feature == "ENSG00000120129")
```

## Student-t shape prior (default)

The default `shape_prior = "student_t"` adds

`shape ~ 1 + offset(log(1/dispersion))`

and places a Student-t prior on the shape intercept. The scale of that prior comes from `dispersion_degrees_freedom` via `sqrt(trigamma(d_eff / 2))`.

```{r one-gene-student, message = FALSE, warning = FALSE, eval = run_vignette}
fit_t <- estimate_gene(
  se_gene,
  formula = ~ dex + (1 | cell),
  offset = "offset",
  dispersion = "dispersion",
  shape_prior = "student_t",
  shape_prior_df = 3,
  family = brms::zero_inflated_negbinomial(),
  chains = 1,
  iter = 250,
  warmup = 100,
  cores = 1,
  backend = "cmdstanr",
  refresh = 0,
  silent = 2
)

hypothesis_gene(fit_t, "dextrt = 0")

adjust_gene(
  fit_t,
  nullify = "dex",
  re_formula = ~(1 | cell)
)
```

## Conjugate gamma shape prior

Alternatively, `shape_prior = "gamma"` puts a gamma prior directly on the scalar `shape` parameter (no shape submodel). The prior mean is `1 / phi_g` for that gene.

```{r one-gene-gamma, message = FALSE, warning = FALSE, eval = run_vignette}
fit_g <- estimate_gene(
  se_gene,
  formula = ~ dex + (1 | cell),
  offset = "offset",
  dispersion = "dispersion",
  shape_prior = "gamma",
  family = brms::zero_inflated_negbinomial(),
  chains = 1,
  iter = 250,
  warmup = 100,
  cores = 1,
  backend = "cmdstanr",
  refresh = 0,
  silent = 2
)
```

Both routes share the same log-scale spread, `trigamma(d_eff / 2)`. They differ in tail weight and in whether they centre the *median* (Student-t) or the *mean* (gamma) of the shape on edgeR's estimate. See `?estimate_dispersion` for the derivation.

# Full scalable pipeline

For many genes, start a pipeline with `brmDE()`, then append `estimate()`, `hypothesis()`, and `adjust()`. Printing the object runs it (`print` calls `evaluate_hpc()`).

`brmDE()` does not take a gene list: offset and dispersion belong on the full matrix. After those columns exist, `filter()` any feature set and pass that object into the pipeline.

```{r pipeline, message = FALSE, warning = FALSE, eval = run_vignette}
# `se` already has offset + dispersion from above; subset genes here
pipeline <- se |>
  filter(.feature == "ENSG00000120129") |>
  brmDE(store = tempfile(pattern = "brmde_")) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    shape_prior = "student_t",
    family = brms::negbinomial(),
    chains = 1,
    iter = 250,
    warmup = 100,
    cores = 1,
    refresh = 0,
    silent = 2
  ) |>
  hypothesis("dextrt = 0") |>
  adjust(nullify = "dex")

# Runs the targets graph
pipeline
```

# What the pipeline writes

| Step | Column / target | Role |
|------|-----------------|------|
| tidybulk / tidySE (or `brmDE()`) | `colData` `"multiplier"`, `"offset"` | TMM scale factor and `log(1 / multiplier)` |
| `estimate()` → `estimate_dispersion()` | `rowData` `"dispersion"`, `"dispersion_degrees_freedom"` | edgeR \(\phi_g\) and \(d_{\mathrm{eff}}\) |
| `estimate()` → `estimate_gene()` | `brms_fit` per gene | posterior fit with shape prior from dispersion |
| `hypothesis()` | hypothesis table | posterior tests |
| `adjust()` | adjusted residuals / fitted | nuisance covariates nullified |

# Dispersion priors at a glance

| `shape_prior` | Model for shape | Prior |
|---------------|-----------------|-------|
| `"student_t"` (default) | `shape ~ 1 + offset(log(1/dispersion))` | Student-t on the intercept; scale from `d_eff` |
| `"gamma"` | scalar `shape` (no submodel) | `gamma(d_eff/2, d_eff * dispersion/2)` |

`prior.df` from `estimateDisp()` is shared across genes (unless `robust = TRUE` in edgeR). What varies per gene is the tagwise `dispersion` that sets the prior location.
