---
title: "Mixed-effect modelling for differential gene expression with brmDE"
author: "Stefano Mangiola"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-effect modelling for differential gene expression with brmDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Fitting chunks run only when CmdStan is available (same gate as sccomp).
run_vignette <- instantiate::stan_cmdstan_exists()
```

# Overview

`brmDE` does mixed-effect modelling for differential gene expression analyses, based on Bayesian regression through [brms](https://paul-buerkner.github.io/brms/). Each gene is fit on its own, then those fits compose into a [tidytargets](https://github.com/stemangiola/tidytargets) `targets` pipeline. The typical flow is:

1. scale with TMM (`tidybulk::scale_abundance()`) and set `offset = log(1 / multiplier)`
2. estimate **dispersion** once on the full matrix (`tidybulk::estimate_dispersion()`)
3. fit each gene with `estimate_gene()`, using that dispersion as a **prior** on the negative binomial shape — the gene's own counts can still pull the posterior away from it
4. optionally run posterior **hypothesis** tests, **adjust** for nuisance covariates, and **plot** a gene across a factor

Step 1 happens outside `brmDE`. The package does no normalisation of its own: `brmDE()` and `estimate_gene()` both expect the offset column to be on the object already, so you stay in control of how library size is handled.

Fitting chunks below are gated by `instantiate::stan_cmdstan_exists()`, so the vignette still builds when CmdStan is missing.

# Install CmdStan

`brmDE` uses `cmdstanr` (not on CRAN) for the default Stan backend. Three steps:

1. R package install of `brmDE`
2. R install of `cmdstanr`
3. `cmdstanr` install of CmdStan itself

```{r install, eval = FALSE}
# from this repository
devtools::install()

install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev/", getOption("repos"))
)

cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

```{r packages, message = FALSE}
library(brmDE)
library(tidybulk)
library(tidySummarizedExperiment)
library(SummarizedExperiment)
data("airway", package = "airway")

airway <- airway |>
  mutate(
    dex = relevel(factor(dex), ref = "untrt"),
    cell = droplevels(cell)
  )
```

# One gene, with dispersion

For a single gene you still prepare the offset and dispersion on a multi-gene matrix. After those columns exist, `filter()` the feature set you want to fit. These steps do not need CmdStan.

Get the TMM `multiplier` from `tidybulk::scale_abundance()`, then form the library-size offset as `log(1 / multiplier)` with `mutate()` from tidySummarizedExperiment.

```{r one-gene-dispersion, message = FALSE}

se <- airway |>
  identify_abundant(formula_design = ~ dex) |>
  scale_abundance(method = "TMM") |>
  # library size factor is the reciprocal of the multiplier
  mutate(offset = log(1 / multiplier))

# Because tidybulk::estimate_dispersion() uses edgeR, which has no mixed
# models, write the fixed-effect analogue yourself:
# `dex + (1|cell)` becomes `dex + cell` and `dex + (dex|cell)` becomes `dex*cell`.
se <- se |> tidybulk::estimate_dispersion(formula_abundance = ~ dex + cell)

# One rowData column per estimate, named by dispersion_column and
# dispersion_degrees_freedom_column (defaults shown):
#   dispersion                 = tagwise phi_g
#   dispersion_degrees_freedom = d_eff = df.residual + prior.df
head(as.data.frame(rowData(se)[, c("dispersion", "dispersion_degrees_freedom")]))
head(se$offset)

```

## Two formulas, and the offsets brmDE adds

`formula_abundance` is the mean model. For the gene-wise fit it can include random effects; the library size offset goes here. For `tidybulk::estimate_dispersion()` you already wrote the fixed-effect analogue that edgeR can use.

`formula_dispersion` is the model for the negative binomial shape, `~1` by default. It is where the edgeR dispersion offset goes.

You write neither offset. Both are appended in the backend, and the assembled formulas are printed so you can confirm what was fitted:

```
Abundance model (offset added by brmDE): counts ~ dex + (1 | cell) + offset(offset)
Dispersion model (offset added by brmDE): shape ~ 1 + offset(log(1/dispersion))
```

The edgeR columns are a **prior**, not a plug-in. `estimate_gene()` centres the negative binomial shape on `1/dispersion` and uses `dispersion_degrees_freedom` to set how tightly; the gene-wise likelihood is then free to move the posterior away from that location. That is the point of bringing a whole-matrix estimate in this way: it informs the fit when the gene is sparse, without locking the shape to edgeR if the counts disagree. How far a gene can diverge is the prior scale (`d_eff`) and the tail of `shape_prior` (Student-t is more permissive than gamma).

`dispersion` and `dispersion_degrees_freedom` both default to `NULL`. Leaving `dispersion` unset puts a zero offset on the shape submodel (the intercept is then `log(shape)` itself), which is useful when the sample size is large enough that an informative location from edgeR is unnecessary. Leaving `dispersion_degrees_freedom` unset gives the Student-t intercept a log-scale SD of 1, so the prior scale does not depend on previous tools. Computing both columns with `tidybulk::estimate_dispersion()` and passing their names, as below, is the preferred starting point.

## Student-t shape prior (default)

The default `shape_prior = "student_t"` builds the shape submodel from `formula_dispersion` and places a Student-t prior on its intercept. With `dispersion_degrees_freedom` passed in, the scale of that prior comes from `sqrt(trigamma(d_eff / 2))`. Because the dispersion enters as an offset on a log link, any terms you add to `formula_dispersion` describe departures from edgeR's per-gene estimate, on top of the intercept's ability to leave that estimate when the data warrant it.

```{r one-gene-student, message = FALSE, warning = FALSE, eval = run_vignette}
fit_t <- se |>
  filter(.feature == "ENSG00000120129") |>
  estimate_gene(
    formula_abundance = ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::zero_inflated_negbinomial(),
    chains = 1,            # set for vignette build
    draws_warmup = 100,    # set for vignette build
    draws_sampling = 150,  # set for vignette build
    cores = 1,             # set for vignette build
    refresh = 0,           # set for vignette build
    silent = 2             # set for vignette build
  )

# Keep the dex main effect; drop the cell random intercept.
adjust_gene(fit_t, re_formula = ~0)
```

## Hypothesis testing

Name the effect on its own and `hypothesis_gene()` asks the question a DE analysis actually wants: is this effect big enough to care about, and which way does it go?

```{r one-gene-hypotheses, message = FALSE, warning = FALSE, eval = run_vignette}
# Is the dex effect bigger than one log2 fold change, and in which direction?
hypothesis_gene(fit_t, "dextrt")
```

`test_above_log2FC` is the size an effect has to beat, in log2 fold change units, so the default `1` asks for a doubling of expression and `2` for a quadrupling. `pH0` is the probability that the better supported side is wrong:

\[
p_{H0} = P(\text{effect too small}) + P(\text{effect runs the other way}) = 1 - \max(p_{+}, p_{-}).
\]

Both ways of being wrong are in that number, which is the point of splitting the sides. Testing `P(|effect| > threshold)` instead would add the two tails together, and a posterior sitting on zero with heavy tails has mass beyond the threshold on *both* sides at once — a large two-sided probability for an effect whose sign is not even determined. Counting the sides separately cannot make that mistake: `pH0` is at least 1/2 for any posterior symmetric about zero, whatever its tails, and climbs to 1 for a posterior lying entirely inside the threshold. Setting `test_above_log2FC = 0` tests only the sign and gives the local false sign rate.

Which way the effect goes is left to `log2_fold_change`. There is no direction label: a gene whose posterior sits entirely inside the threshold supports no direction at all, so labelling one would be inventing a claim the fit does not make.

Because `pH0` is a probability rather than a test statistic, a false discovery rate follows by averaging: the expected number of false discoveries in a set of genes is the sum of their null probabilities, so the mean `pH0` over the genes you selected *is* the expected proportion of false discoveries among them. `false_discovery_rate()` sorts genes by `pH0` and reports that running mean, and pipelines add it as an `fdr` column. No null distribution and no p-values are involved, and since a cumulative mean of an ascending sequence is already monotone, no Benjamini-Hochberg step-up is needed either.

One limit worth keeping in mind: `pH0` is counted from draws, so it cannot resolve below `1 / ndraws`. A `pH0` of 0 means "smaller than this fit can measure", and an `fdr` far below `1 / ndraws` is not supported by the draws behind it. The fits in this vignette are deliberately short in order to build quickly.

Writing the effect against zero instead, `hypothesis_gene(fit_t, "dextrt = 0")`, fills `pH0` from a Savage-Dickey Bayes factor rather than by counting draws. The effect size columns do not move, but the number now needs prior draws (`sample_prior = "yes"` in `estimate_gene()`) and a proper prior on the parameter, and it is `NA` without them. The `pH0_from` column records which of the two a row used, `"posterior_draws"` or `"bayes_factor"`, because a `pH0` derived from a Bayes factor carries assumptions the counted one does not. See `?hypothesis_gene` for the full comparison.

## Group differences

The same test applies to random intercepts. Airway has four cell lines; asking whether each cell's intercept exceeds one log2 fold change from the population mean is four hypotheses. `class = "r"` tells brms to look among the group-level coefficients.

```{r one-gene-cells, message = FALSE, warning = FALSE, eval = run_vignette}
hypothesis_gene(
  fit_t,
  c(
    "`cell[N052611,Intercept]`",
    "`cell[N061011,Intercept]`",
    "`cell[N080611,Intercept]`",
    "`cell[N61311,Intercept]`"
  ),
  class = "r"
)
```

## Plotting the fit

The same fit is worth looking at. `plot_boxplot()` draws the TMM-normalised counts of one gene across a factor; the construction is the same as `sccomp`'s boxplot. Black boxplots and points are the counts, blue boxplots are draws from the posterior predictive distribution. If the model is descriptively adequate, the blue boxplots should roughly overlay the black ones.

The library size is always taken out, and through the model rather than by dividing the counts: `adjust_gene()` predicts with the offset zeroed, which for the TMM offset above leaves the normalised abundance.

`ENSG00000120129` is DUSP1, a dexamethasone-responsive gene. The hypothesis above asked whether its dex effect clears one log2 fold change; the plot is that effect in the counts.

```{r one-gene-plot, message = FALSE, warning = FALSE, eval = run_vignette, fig.width = 6, fig.height = 4}
plot_boxplot(fit_t, factor = "dex", feature = "ENSG00000120129")
```

Those counts still carry the cell-line intercepts, so for a mixed model the dex contrast can be hard to see in one dimension. `remove_unwanted_effects = TRUE` takes the random intercepts out too, keeping `dex` alone, and redraws the posterior predictive from that same reduced model.

```{r one-gene-plot-adjusted, message = FALSE, warning = FALSE, eval = run_vignette, fig.width = 6, fig.height = 4}
plot_boxplot(
  fit_t,
  factor = "dex",
  feature = "ENSG00000120129",
  remove_unwanted_effects = TRUE
)
```

`plot_boxplot()` returns a ggplot object, so further layers can be added as usual.

# Full scalable pipeline

For many genes, start a pipeline with `brmDE()`, then append `estimate()`, `hypothesis()`, and `adjust()`. Printing the object runs it (`print` calls `tt_evaluate()`).

The pipeline fits genes and nothing else. Offset and dispersion are whole-matrix quantities, so they are computed above, before any gene is chosen; once those columns exist you can `filter()` any feature set and pass that object in. Five genes are enough to show the gene-wise branches that `estimate()` maps over; pass a local `crew` controller to `brmDE()` so those branches run in parallel on your machine (the default is sequential).

```{r pipeline, message = FALSE, warning = FALSE, eval = run_vignette}
library(crew)
store <- tempfile(pattern = "brmde_")
pipeline <- se |>
  filter(.feature %in% c(
    "ENSG00000120129",
    "ENSG00000000003",
    "ENSG00000000005",
    "ENSG00000000419",
    "ENSG00000000457"
  )) |>
  brmDE(
    store = store,
    computing_resources = crew_controller_local(workers = 5)
  ) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::zero_inflated_negbinomial(),
    chains = 1,            # set for vignette build
    draws_warmup = 100,    # set for vignette build
    draws_sampling = 150,  # set for vignette build
    cores = 1,             # set for vignette build
    refresh = 0,           # set for vignette build
    silent = 2             # set for vignette build
  ) |>
  hypothesis("dextrt") |>
  adjust(re_formula = ~0)

# Runs the targets graph
pipeline
```

## What comes back, and where the fits are

One row per gene and contrast. `estimate()` contributes the gene column; `hypothesis()` adds the columns that make the run readable, unnested into the result rather than left in a list column, and `adjust()` its own table. Each row carries `log2_fold_change` and `pH0` for that contrast, the `fdr` those `pH0` values give across genes, and the convergence of that contrast's own draws: `rhat`, `ess_bulk`, and `mcse`, the Monte Carlo error of the reported `estimate`, which you can read against the width of the interval beside it. Those are diagnostics of the quantity being reported rather than of the parameters behind it, which is the distinction that matters in practice: a `cell` level that drifts between chains cancels in a difference of levels, so the contrast is often trustworthy when its parts are not. `ess_bulk` also bounds how far `pH0` can be trusted, since a probability counted from draws cannot resolve below `1 / ndraws` and correlated draws are worth fewer than their number. That denominator is not in the table, because it is a property of the fits rather than of any row; `estimate()` records it on the pipeline instead, which is what `plot_volcano()` reads it from below.

The fits themselves are not in the result. A `brmsfit` carries its draws and its data, so twenty thousand of them cannot be held in one object, and at that scale the statistics are what you want anyway: they tell you which handful of genes to open. Those stay in the targets store; `targets::tar_read(brms_fit, branches = i, store = store)` reads branch `i`. With `bundle = 1` that is the gene's row in the result. The gene id is on the result, not on the fit: pass `feature = result$.feature[[i]]` to `plot_boxplot()`.

```{r diagnostics, eval = run_vignette}
result <- pipeline |> tt_evaluate()
result

# Open the gene whose contrast mixed worst. The gene id is on the result.
worst <- which.max(result$rhat)
targets::tar_read(brms_fit, branches = worst, store = store)
# plot_boxplot(..., factor = "dex", feature = result$.feature[[worst]])
```

## Plotting the run

`plot_volcano()` draws that run as a volcano: `log2_fold_change` against `pH0` on a reversed log scale, so the genes a run is about sit at the top. Genes below `significance_threshold` (default `0.05`) are drawn red and larger. The shape is the `tidybulk` volcano.

The one addition is the shaded band along the bottom. `pH0` is counted from draws, so it cannot resolve below `1 / ndraws`, and at transcriptome scale a great many genes come back as exactly 0 — which a log scale has nowhere to put. Those genes are drawn in a band one decade below the dashed resolution line and jittered across it. The band means "smaller than these draws can measure", and the spread inside it is there only to separate the points: it is not a probability, and the axis is deliberately left unlabelled there.

Note that it takes the pipeline, not the result. That is the reason: the table cannot say how many draws are behind its own `pH0`, since the fits stay in the store. The pipeline can. `estimate()` records what it asked its fits for on the object with `tt_metadata()` — the formulas, the column names, and the sampling settings, defaults included — so `plot_volcano()` reads the table out of the store and takes the two entries it needs off the object: `chains * draws_sampling`, here `1 * 150` draws. `tt_metadata(pipeline)` prints the lot. Asking for the plot evaluates the pipeline, exactly as printing it does, so it can also be the first thing you call; targets skips what it already has, and here everything is already built.

```{r volcano, message = FALSE, warning = FALSE, eval = run_vignette, fig.width = 6, fig.height = 4}
pipeline |> plot_volcano()
```

Five genes make a thin volcano; the band earns its place over a whole transcriptome. `plot_volcano()` returns a ggplot object, so a run holding several contrasts can be split with `+ facet_wrap(~ hypothesis)`, as the `tidybulk` volcano facets by method.

## SLURM

On a cluster, pass a `crew.cluster` controller to `brmDE()` as `computing_resources`. Gene-wise targets then launch as SLURM jobs (run this block from a login node, not in the vignette build).

```{r slurm, eval = FALSE}
library(crew.cluster)

pipeline <- se |>
  filter(.feature %in% c(
    "ENSG00000120129",
    "ENSG00000000003",
    "ENSG00000000005",
    "ENSG00000000419",
    "ENSG00000000457"
  )) |>
  brmDE(
    store = "/path/to/brmde_store",
    computing_resources = crew_controller_slurm(
      workers = 100,
      options_cluster = crew_options_slurm(
        cpus_per_task = 2,
        memory_gigabytes_per_cpu = 5
      )
    )
  ) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::negbinomial()
  ) |>
  hypothesis("dextrt") |>
  adjust(re_formula = ~0)

pipeline
```

## Bundling genes into fewer jobs

At transcriptome scale, if the sample size is small and each gene estimation is quick, the scheduler must handle tens of thousands of jobs whose queueing cost can rival the fits themselves. `bundle` is how many genes `estimate()` puts in each fit target, fitted in sequence; the default `10` is ten genes per target, and `1` is one target per gene. `hypothesis()` and `adjust()` map over the fit target, so they inherit the coarser branching without being told about it, and the result is still one row per gene in the same order.
On the other hand, if the sample size is large and each gene estimation is slow, you should not bundle as you risk less that the failure of one gene would compromise the whole bundle.

```{r bundle, eval = FALSE}
pipeline <- se |>
  brmDE(
    store = "/path/to/brmde_store",
    computing_resources = crew_controller_slurm(workers = 100)
  ) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    bundle = 100    # 20,000 genes become 200 jobs of 100 fits each
  ) |>
  hypothesis("dextrt")
```

Pick `bundle` against a single job rather than against the gene count, because that is exactly what it fixes: a job takes about `bundle` times as long as one fit and lands on disk as a single `.rds` holding `bundle` `brmsfit` objects, whether you are running 200 genes or 20,000. Bundling also coarsens `targets`' bookkeeping. A gene that fails takes its bundle with it, and changing the gene set reshuffles which genes share a bundle, so everything refits. Leave `bundle` at `1` when you are iterating on a gene set and relying on incremental reruns.

# What the pipeline writes

| Step | Column / target | Role |
|------|-----------------|------|
| tidybulk / tidySE, before `brmDE()` | `colData` `"multiplier"`, `"offset"` | TMM scale factor and `log(1 / multiplier)` |
| `tidybulk::estimate_dispersion()`, before `brmDE()` | `rowData` `"dispersion"`, `"dispersion_degrees_freedom"` | edgeR \(\phi_g\) and \(d_{\mathrm{eff}}\) |
| `estimate()` → `estimate_gene()` | `brms_fit` per gene, in the store | posterior fit; edgeR dispersion is the shape prior, not a fixed value |
| `hypothesis()` | hypothesis table | `log2_fold_change` and `pH0` per gene, `rhat`/`ess_bulk`/`mcse` per contrast, plus `fdr` across genes |
| `adjust()` | adjusted residuals / fitted | dex kept; random intercepts dropped |
| `plot_boxplot()` | ggplot of one `brmsfit` | observed or adjusted counts across a factor, blue posterior predictive overlay |
| `plot_volcano()` | ggplot of the pipeline | `log2_fold_change` against `pH0`, with everything below `1 / ndraws` jittered in a shaded band |

# Dispersion priors at a glance

| `shape_prior` | Model for shape | Prior |
|---------------|-----------------|-------|
| `"student_t"` (default) | `shape ~ 1 + offset(log(1/dispersion))`, or `offset(0)` if `dispersion` is `NULL` | Student-t on the intercept; scale from `d_eff`, or log-scale SD 1 if `dispersion_degrees_freedom` is `NULL` |
| `"gamma"` | scalar `shape` (no submodel) | `gamma(d_eff/2, d_eff * dispersion/2)` (needs both columns) |

`prior.df` from `estimateDisp()` is shared across genes (unless `robust = TRUE` in edgeR). What varies per gene is the tagwise `dispersion` that sets the prior location. In either form the gene-wise counts update that prior: the posterior shape is not forced to equal `1/dispersion`.

# Abundance priors at a glance

The default prior set applies whenever `prior` is left `NULL` and `family` is a negative binomial, zero-inflated or not.

| Parameter | Prior | Set by |
|-----------|-------|--------|
| `Intercept` | `student_t(3, mean(log1p(counts / exp(offset))), 1.5)` | the gene's own data |
| `b` (coefficients) | `student_t(coefficient_prior_df, 0, coefficient_prior_scale)`, `student_t(3, 0, 0.7)` by default, one log2 fold change | `coefficient_prior_scale`, `coefficient_prior_df` |
| `Intercept` of `shape` | `student_t(shape_prior_df, 0, scale)` | `dispersion_degrees_freedom`, `shape_prior_df` |
| `shape` (scalar) | `gamma(d_eff/2, d_eff * dispersion/2)` | `shape_prior = "gamma"` |

Passing your own `prior` replaces the whole set.
