---
title: "Gene-wise Bayesian DE with brmDE"
author: "Stefano Mangiola"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Gene-wise Bayesian DE with brmDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Fitting chunks run only when CmdStan is available (same gate as sccomp).
run_vignette <- instantiate::stan_cmdstan_exists()
```

# Overview

`brmDE` fits one gene at a time with [brms](https://paul-buerkner.github.io/brms/), then composes those fits into an [HPCell](https://github.com/MangiolaLaboratory/HPCell) `targets` pipeline. The typical flow is:

1. scale with TMM (`tidybulk::scale_abundance()`) and set `offset = log(1 / multiplier)`
2. estimate edgeR **dispersion** once on the full matrix (`estimate_dispersion()`)
3. fit each gene with `estimate_gene()`, using the dispersion to inform the negative binomial shape prior
4. optionally run posterior **hypothesis** tests and **adjust** for nuisance covariates

Step 1 happens outside `brmDE`. The package does no normalisation of its own: `brmDE()` and `estimate_gene()` both expect the offset column to be on the object already, so you stay in control of how library size is handled.

Fitting chunks below are gated by `instantiate::stan_cmdstan_exists()`, so the vignette still builds when CmdStan is missing.

# Install CmdStan

`brmDE` uses `cmdstanr` (not on CRAN) for the default Stan backend. Three steps:

1. R package install of `brmDE`
2. R install of `cmdstanr`
3. `cmdstanr` install of CmdStan itself

```{r install, eval = FALSE}
# from this repository
devtools::install()

install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev/", getOption("repos"))
)

cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

```{r packages, message = FALSE}
library(brmDE)
library(tidybulk)
library(tidySummarizedExperiment)
library(SummarizedExperiment)
data("airway", package = "airway")

airway$dex <- relevel(factor(airway$dex), ref = "untrt")
airway$cell <- droplevels(airway$cell)
```

# One gene, with dispersion

For a single gene you still prepare the offset and dispersion on a multi-gene matrix. After those columns exist, `filter()` the feature set you want to fit. These steps do not need CmdStan.

Get the TMM `multiplier` from `tidybulk::scale_abundance()`, then form the library-size offset as `log(1 / multiplier)` with `mutate()` from tidySummarizedExperiment (same convention as the immuneBodyMap / HPCell pipeline).

```{r one-gene-dispersion, message = FALSE}

se <- airway |>
  identify_abundant(formula_design = ~ dex) |>
  scale_abundance(method = "TMM") |>
  # library size factor is the reciprocal of the multiplier
  mutate(offset = log(1 / multiplier))

# Because the edgeR dispersion estimation (in the backend of the `estimate_dispersion` function)
# cannot model mixed effect model, we do the next best thing
# by adding the cell random effect to the formula_abundance as fixed effect
# So `dex + (1|cell)` becomes `dex + cell` and `dex + (dex|cell)` becomes  `dex*cell`
se <- se |> estimate_dispersion(formula_abundance = ~ dex + cell)

# One rowData column per estimate, both named by arguments of the same name:
#   dispersion                 = tagwise phi_g
#   dispersion_degrees_freedom = d_eff = df.residual + prior.df
head(as.data.frame(rowData(se)[, c("dispersion", "dispersion_degrees_freedom")]))
head(se$offset)

```

## Two formulas, and the offsets brmDE adds

`formula_abundance` is the mean model. It is the same formula `estimate_dispersion()` gave to edgeR above, and it is where the library size offset goes.

`formula_dispersion` is the model for the negative binomial shape, `~1` by default. It is where the edgeR dispersion offset goes.

You write neither offset. Both are appended in the backend, and the assembled formulas are printed so you can confirm what was fitted:

```
Abundance model (offset added by brmDE): counts ~ dex + (1 | cell) + offset(offset)
Dispersion model (offset added by brmDE): shape ~ 1 + offset(log(1/dispersion))
```

`dispersion` and `dispersion_degrees_freedom` both default to `NULL`. Leaving `dispersion` unset puts a zero offset on the shape submodel (the intercept is then `log(shape)` itself), which is useful when the sample size is large enough that an informative location from edgeR is unnecessary. Leaving `dispersion_degrees_freedom` unset gives the Student-t intercept a log-scale SD of 1, so the prior scale does not depend on previous tools. Computing both columns with `estimate_dispersion()` and passing their names, as below, is the preferred starting point.

## Student-t shape prior (default)

The default `shape_prior = "student_t"` builds the shape submodel from `formula_dispersion` and places a Student-t prior on its intercept. With `dispersion_degrees_freedom` passed in, the scale of that prior comes from `sqrt(trigamma(d_eff / 2))`. Because the dispersion enters as an offset on a log link, any terms you add to `formula_dispersion` describe departures from edgeR's per-gene estimate.

```{r one-gene-student, message = FALSE, warning = FALSE, eval = run_vignette}
fit_t <- se |>
  filter(.feature == "ENSG00000120129") |>
  estimate_gene(
    formula_abundance = ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::zero_inflated_negbinomial(),
    chains = 1,       # set for vignette build
    iter = 250,       # set for vignette build
    warmup = 100,     # set for vignette build
    cores = 1,        # set for vignette build
    refresh = 0,      # set for vignette build
    silent = 2        # set for vignette build
  )

hypothesis_gene(fit_t, "dextrt = 0")

# Keep the dex main effect; drop the cell random intercept.
adjust_gene(fit_t, re_formula = ~0)
```

# Full scalable pipeline

For many genes, start a pipeline with `brmDE()`, then append `estimate()`, `hypothesis()`, and `adjust()`. Printing the object runs it (`print` calls `evaluate_hpc()`).

The pipeline fits genes and nothing else. Offset and dispersion are whole-matrix quantities, so they are computed above, before any gene is chosen; once those columns exist you can `filter()` any feature set and pass that object in. Five genes are enough to show the gene-wise branches that `estimate()` maps over; pass a local `crew` controller to `brmDE()` so those branches run in parallel on your machine (the default is sequential).

```{r pipeline, message = FALSE, warning = FALSE, eval = run_vignette}
library(crew)
pipeline <- se |>
  filter(.feature %in% c(
    "ENSG00000120129",
    "ENSG00000000003",
    "ENSG00000000005",
    "ENSG00000000419",
    "ENSG00000000457"
  )) |>
  brmDE(computing_resources = crew_controller_local(workers = 5)) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::zero_inflated_negbinomial(),
    chains = 1,       # set for vignette build
    iter = 250,       # set for vignette build
    warmup = 100,     # set for vignette build
    cores = 1,        # set for vignette build
    refresh = 0,      # set for vignette build
    silent = 2        # set for vignette build
  ) |>
  hypothesis("dextrt = 0") |>
  adjust(re_formula = ~0)

# Runs the targets graph
pipeline
```

## SLURM

On a cluster, pass a `crew.cluster` controller to `brmDE()` as `computing_resources`. Gene-wise targets then launch as SLURM jobs (run this block from a login node, not in the vignette build).

```{r slurm, eval = FALSE}
library(crew.cluster)

pipeline <- se |>
  filter(.feature %in% c(
    "ENSG00000120129",
    "ENSG00000000003",
    "ENSG00000000005",
    "ENSG00000000419",
    "ENSG00000000457"
  )) |>
  brmDE(
    store = "/path/to/brmde_store",
    computing_resources = crew_controller_slurm(
      slurm_memory_gigabytes_per_cpu = 5,
      workers = 100,
      tasks_max = 5
    )
  ) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::negbinomial()
  ) |>
  hypothesis("dextrt = 0") |>
  adjust(re_formula = ~0)

pipeline
```

# What the pipeline writes

| Step | Column / target | Role |
|------|-----------------|------|
| tidybulk / tidySE, before `brmDE()` | `colData` `"multiplier"`, `"offset"` | TMM scale factor and `log(1 / multiplier)` |
| `estimate_dispersion()`, before `brmDE()` | `rowData` `"dispersion"`, `"dispersion_degrees_freedom"` | edgeR \(\phi_g\) and \(d_{\mathrm{eff}}\) |
| `estimate()` → `estimate_gene()` | `brms_fit` per gene | posterior fit with shape prior from dispersion |
| `hypothesis()` | hypothesis table | posterior tests |
| `adjust()` | adjusted residuals / fitted | dex kept; random intercepts dropped |

# Dispersion priors at a glance

| `shape_prior` | Model for shape | Prior |
|---------------|-----------------|-------|
| `"student_t"` (default) | `shape ~ 1 + offset(log(1/dispersion))`, or `offset(0)` if `dispersion` is `NULL` | Student-t on the intercept; scale from `d_eff`, or log-scale SD 1 if `dispersion_degrees_freedom` is `NULL` |
| `"gamma"` | scalar `shape` (no submodel) | `gamma(d_eff/2, d_eff * dispersion/2)` (needs both columns) |

`prior.df` from `estimateDisp()` is shared across genes (unless `robust = TRUE` in edgeR). What varies per gene is the tagwise `dispersion` that sets the prior location.
