---
title: "Mixed-effect modelling for differential gene expression with brmDE"
author: "Stefano Mangiola"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-effect modelling for differential gene expression with brmDE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
# Fitting chunks run only when CmdStan is available (same gate as sccomp).
run_vignette <- instantiate::stan_cmdstan_exists()
```

# Overview

`brmDE` does mixed-effect modelling for differential gene expression analyses, based on Bayesian regression through [brms](https://paul-buerkner.github.io/brms/). Each gene is fit on its own, then those fits compose into a [tidytargets](https://github.com/stemangiola/tidytargets) `targets` pipeline. The typical flow is:

1. scale with TMM (`tidybulk::scale_abundance()`) and set `offset = log(1 / multiplier)`
2. estimate edgeR **dispersion** once on the full matrix (`estimate_dispersion()`)
3. fit each gene with `estimate_gene()`, using the dispersion to inform the negative binomial shape prior
4. optionally run posterior **hypothesis** tests and **adjust** for nuisance covariates

Step 1 happens outside `brmDE`. The package does no normalisation of its own: `brmDE()` and `estimate_gene()` both expect the offset column to be on the object already, so you stay in control of how library size is handled.

Fitting chunks below are gated by `instantiate::stan_cmdstan_exists()`, so the vignette still builds when CmdStan is missing.

# Install CmdStan

`brmDE` uses `cmdstanr` (not on CRAN) for the default Stan backend. Three steps:

1. R package install of `brmDE`
2. R install of `cmdstanr`
3. `cmdstanr` install of CmdStan itself

```{r install, eval = FALSE}
# from this repository
devtools::install()

install.packages(
  "cmdstanr",
  repos = c("https://stan-dev.r-universe.dev/", getOption("repos"))
)

cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

```{r packages, message = FALSE}
library(brmDE)
library(tidybulk)
library(tidySummarizedExperiment)
library(SummarizedExperiment)
data("airway", package = "airway")

airway <- airway |>
  mutate(
    dex = relevel(factor(dex), ref = "untrt"),
    cell = droplevels(cell)
  )
```

# One gene, with dispersion

For a single gene you still prepare the offset and dispersion on a multi-gene matrix. After those columns exist, `filter()` the feature set you want to fit. These steps do not need CmdStan.

Get the TMM `multiplier` from `tidybulk::scale_abundance()`, then form the library-size offset as `log(1 / multiplier)` with `mutate()` from tidySummarizedExperiment.

```{r one-gene-dispersion, message = FALSE}

se <- airway |>
  identify_abundant(formula_design = ~ dex) |>
  scale_abundance(method = "TMM") |>
  # library size factor is the reciprocal of the multiplier
  mutate(offset = log(1 / multiplier))

# Because the edgeR dispersion estimation (in the backend of the `estimate_dispersion` function)
# cannot model mixed effect model, we do the next best thing
# by adding the cell random effect to the formula_abundance as fixed effect
# So `dex + (1|cell)` becomes `dex + cell` and `dex + (dex|cell)` becomes  `dex*cell`
se <- se |> estimate_dispersion(formula_abundance = ~ dex + cell)

# One rowData column per estimate, both named by arguments of the same name:
#   dispersion                 = tagwise phi_g
#   dispersion_degrees_freedom = d_eff = df.residual + prior.df
head(as.data.frame(rowData(se)[, c("dispersion", "dispersion_degrees_freedom")]))
head(se$offset)

```

## Two formulas, and the offsets brmDE adds

`formula_abundance` is the mean model. It is the same formula `estimate_dispersion()` gave to edgeR above, and it is where the library size offset goes.

`formula_dispersion` is the model for the negative binomial shape, `~1` by default. It is where the edgeR dispersion offset goes.

You write neither offset. Both are appended in the backend, and the assembled formulas are printed so you can confirm what was fitted:

```
Abundance model (offset added by brmDE): counts ~ dex + (1 | cell) + offset(offset)
Dispersion model (offset added by brmDE): shape ~ 1 + offset(log(1/dispersion))
```

`dispersion` and `dispersion_degrees_freedom` both default to `NULL`. Leaving `dispersion` unset puts a zero offset on the shape submodel (the intercept is then `log(shape)` itself), which is useful when the sample size is large enough that an informative location from edgeR is unnecessary. Leaving `dispersion_degrees_freedom` unset gives the Student-t intercept a log-scale SD of 1, so the prior scale does not depend on previous tools. Computing both columns with `estimate_dispersion()` and passing their names, as below, is the preferred starting point.

## Student-t shape prior (default)

The default `shape_prior = "student_t"` builds the shape submodel from `formula_dispersion` and places a Student-t prior on its intercept. With `dispersion_degrees_freedom` passed in, the scale of that prior comes from `sqrt(trigamma(d_eff / 2))`. Because the dispersion enters as an offset on a log link, any terms you add to `formula_dispersion` describe departures from edgeR's per-gene estimate.

```{r one-gene-student, message = FALSE, warning = FALSE, eval = run_vignette}
fit_t <- se |>
  filter(.feature == "ENSG00000120129") |>
  estimate_gene(
    formula_abundance = ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::zero_inflated_negbinomial(),
    chains = 1,       # set for vignette build
    iter = 250,       # set for vignette build
    warmup = 100,     # set for vignette build
    cores = 1,        # set for vignette build
    refresh = 0,      # set for vignette build
    silent = 2        # set for vignette build
  )

# Keep the dex main effect; drop the cell random intercept.
adjust_gene(fit_t, re_formula = ~0)
```

## Point and one-sided hypotheses

The form of the equation you give `hypothesis_gene()` decides how its `post_prob` and `evid_ratio` columns are obtained, and the two routes ask different things of the fit.

A one-sided equation, `"dextrt > 0"`, is a tally: `post_prob` is the fraction of posterior draws with a positive coefficient, and `evid_ratio` is that fraction over its complement. Nothing but the posterior is involved.

A point equation, `"dextrt = 0"`, is a Bayes factor computed by the Savage-Dickey density ratio: the posterior density at 0 divided by the prior density at 0. That needs draws from the prior as well as the posterior, so `estimate_gene()` passes `sample_prior = "yes"` to brms by default. Those draws are taken in Stan's generated quantities block, so they cost no sampling time, and they are what keeps the two columns from being `NA`. Every parameter under test also needs a prior that integrates to one, because brms cannot draw from a flat prior; this is why the default prior set is applied to plain `brms::negbinomial()` fits and not only to zero-inflated ones.

If you are arriving from edgeR, this is the Bayesian counterpart of the likelihood-ratio test that `glmLRT()` gives you for the same coefficient. Both weigh a model with `dex` free against the same model with it held at 0, which is why both want the null to be a restriction of the alternative. The difference is what happens to the parameters: the likelihood-ratio test maximises the likelihood under each hypothesis and refers the ratio to a chi-square distribution to get a p-value, whereas the Bayes factor averages the likelihood over the prior and so needs no null distribution at all. `evid_ratio` is the analogue of the test statistic, with 1 meaning no evidence either way, and `post_prob` is its normalised form. Neither is a p-value, and nothing behind them controls a type-I error rate, so they are not to be thresholded at 0.05.

Two properties of that ratio matter before you read a `post_prob` off a point test.

**It depends on the prior scale, not only on the data.** Widening the prior on a coefficient thins prior mass at 0 and so moves the evidence towards the null however clear the data are. This is Lindley's paradox, and it is why `coefficient_prior_scale` defaults to `0.7`. A coefficient here sits behind a log link, so it is a natural-log fold change, and `0.7` is one log2 fold change: a doubling of expression, which is the order of the effects a DE analysis looks for. Multiples follow in log2 units, `1.4` being two log2 fold changes. Widen it deliberately when your effects are larger, and expect point tests to grow more conservative as you do.

**It is a density estimate, so it needs draws.** The estimate and the interval stabilise long before a density at a single point does, so a short fit gives a usable `ci_lower`/`ci_upper` alongside a `post_prob` that moves from run to run. Around a thousand post-warmup draws is where it starts to hold still. The fits in this vignette are deliberately far shorter than that, in order to build quickly, so read their `post_prob` as illustration rather than as a result.

Group-level coefficients are the one case no refit fixes. brms cannot sample their priors, so a point test on `class = "r"`, which is what `hypothesis = "random_vs_rest"` builds, reports `estimate` and the interval and leaves `post_prob` as `NA`.

```{r one-gene-hypotheses, message = FALSE, warning = FALSE, eval = run_vignette}
# Savage-Dickey Bayes factor against the point null.
hypothesis_gene(fit_t, "dextrt = 0")

# Posterior probability of a positive effect, from the posterior alone.
hypothesis_gene(fit_t, "dextrt > 0")
```

# Full scalable pipeline

For many genes, start a pipeline with `brmDE()`, then append `estimate()`, `hypothesis()`, and `adjust()`. Printing the object runs it (`print` calls `tt_evaluate()`).

The pipeline fits genes and nothing else. Offset and dispersion are whole-matrix quantities, so they are computed above, before any gene is chosen; once those columns exist you can `filter()` any feature set and pass that object in. Five genes are enough to show the gene-wise branches that `estimate()` maps over; pass a local `crew` controller to `brmDE()` so those branches run in parallel on your machine (the default is sequential).

```{r pipeline, message = FALSE, warning = FALSE, eval = run_vignette}
library(crew)
pipeline <- se |>
  filter(.feature %in% c(
    "ENSG00000120129",
    "ENSG00000000003",
    "ENSG00000000005",
    "ENSG00000000419",
    "ENSG00000000457"
  )) |>
  brmDE(computing_resources = crew_controller_local(workers = 5)) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::zero_inflated_negbinomial(),
    chains = 1,       # set for vignette build
    iter = 250,       # set for vignette build
    warmup = 100,     # set for vignette build
    cores = 1,        # set for vignette build
    refresh = 0,      # set for vignette build
    silent = 2        # set for vignette build
  ) |>
  hypothesis("dextrt = 0") |>
  adjust(re_formula = ~0)

# Runs the targets graph
pipeline
```

## SLURM

On a cluster, pass a `crew.cluster` controller to `brmDE()` as `computing_resources`. Gene-wise targets then launch as SLURM jobs (run this block from a login node, not in the vignette build).

```{r slurm, eval = FALSE}
library(crew.cluster)

pipeline <- se |>
  filter(.feature %in% c(
    "ENSG00000120129",
    "ENSG00000000003",
    "ENSG00000000005",
    "ENSG00000000419",
    "ENSG00000000457"
  )) |>
  brmDE(
    store = "/path/to/brmde_store",
    computing_resources = crew_controller_slurm(
      workers = 100,
      options_cluster = crew_options_slurm(
        memory_gigabytes_per_cpu = 5
      )
    )
  ) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    family = brms::negbinomial()
  ) |>
  hypothesis("dextrt = 0") |>
  adjust(re_formula = ~0)

pipeline
```

## Bundling genes into fewer jobs

At transcriptome scale, if the sample size is small and each gene estimation is quick, the scheduler must handle tens of thousands of jobs whose queueing cost can rival the fits themselves. `bundle` is how many genes `estimate()` puts in each fit target, fitted in sequence; `1`, the default, is one target per gene. `hypothesis()` and `adjust()` map over the fit target, so they inherit the coarser branching without being told about it, and the result is still one row per gene in the same order.
On the other hand, if the sample size is large and each gene estimation is slow, you should not bundle as you risk less that the failure of one gene would compromise the whole bundle.

```{r bundle, eval = FALSE}
pipeline <- se |>
  brmDE(
    store = "/path/to/brmde_store",
    computing_resources = crew_controller_slurm(workers = 100)
  ) |>
  estimate(
    ~ dex + (1 | cell),
    offset = "offset",
    dispersion = "dispersion",
    dispersion_degrees_freedom = "dispersion_degrees_freedom",
    bundle = 100    # 20,000 genes become 200 jobs of 100 fits each
  ) |>
  hypothesis("dextrt = 0")
```

Pick `bundle` against a single job rather than against the gene count, because that is exactly what it fixes: a job takes about `bundle` times as long as one fit and lands on disk as a single `.rds` holding `bundle` `brmsfit` objects, whether you are running 200 genes or 20,000. Bundling also coarsens `targets`' bookkeeping. A gene that fails takes its bundle with it, and changing the gene set reshuffles which genes share a bundle, so everything refits. Leave `bundle` at `1` when you are iterating on a gene set and relying on incremental reruns.

# What the pipeline writes

| Step | Column / target | Role |
|------|-----------------|------|
| tidybulk / tidySE, before `brmDE()` | `colData` `"multiplier"`, `"offset"` | TMM scale factor and `log(1 / multiplier)` |
| `estimate_dispersion()`, before `brmDE()` | `rowData` `"dispersion"`, `"dispersion_degrees_freedom"` | edgeR \(\phi_g\) and \(d_{\mathrm{eff}}\) |
| `estimate()` → `estimate_gene()` | `brms_fit` per gene | posterior fit with shape prior from dispersion |
| `hypothesis()` | hypothesis table | posterior tests |
| `adjust()` | adjusted residuals / fitted | dex kept; random intercepts dropped |

# Dispersion priors at a glance

| `shape_prior` | Model for shape | Prior |
|---------------|-----------------|-------|
| `"student_t"` (default) | `shape ~ 1 + offset(log(1/dispersion))`, or `offset(0)` if `dispersion` is `NULL` | Student-t on the intercept; scale from `d_eff`, or log-scale SD 1 if `dispersion_degrees_freedom` is `NULL` |
| `"gamma"` | scalar `shape` (no submodel) | `gamma(d_eff/2, d_eff * dispersion/2)` (needs both columns) |

`prior.df` from `estimateDisp()` is shared across genes (unless `robust = TRUE` in edgeR). What varies per gene is the tagwise `dispersion` that sets the prior location.

# Abundance priors at a glance

The default prior set applies whenever `prior` is left `NULL` and `family` is a negative binomial, zero-inflated or not. All of it is proper, which is what makes a point hypothesis testable.

| Parameter | Prior | Set by |
|-----------|-------|--------|
| `Intercept` | `student_t(3, mean(log1p(counts / exp(offset))), 1.5)` | the gene's own data |
| `b` (coefficients) | `student_t(coefficient_prior_df, 0, coefficient_prior_scale)`, `student_t(3, 0, 0.7)` by default, one log2 fold change | `coefficient_prior_scale`, `coefficient_prior_df` |
| `Intercept` of `shape` | `student_t(shape_prior_df, 0, scale)` | `dispersion_degrees_freedom`, `shape_prior_df` |
| `shape` (scalar) | `gamma(d_eff/2, d_eff * dispersion/2)` | `shape_prior = "gamma"` |

Passing your own `prior` replaces the whole set, so give every parameter you intend to test a proper prior of its own.
