---
title: "Integrating fast frequentist empirical-Bayes gene-expression dispersion estimation as a prior"
author: "Stefano Mangiola"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating fast frequentist empirical-Bayes gene-expression dispersion estimation as a prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.8,
  eval = FALSE
)
run_vignette <-
  requireNamespace("airway", quietly = TRUE) &&
  requireNamespace("edgeR", quietly = TRUE) &&
  requireNamespace("SummarizedExperiment", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
```

# Introduction on `brmDE`

`brmDE` is a flexible tool capable of 

- mixed-effect models

- non-parametric continuous smooth/spline models

- alternative distributions such as NB and ZINB, 

`brmDE` fits one gene at a time to prioritise parallelisation. As a consequence, a single-gene likelihood cannot borrow strength across features, which is exactly what makes empirical-Bayes estimation of feature-wise negative-binomial dispersion useful when the number of replicates is small compared with the parameter size.

Frequentist methods such as `edgeR` and `DESeq2` estimate gene-wise dispersion through empirical Bayes very efficiently, although generally within more constrained fixed-effect model structures.

The key idea is to establish the dispersion prior for `brmDE` cheaply with empirical Bayes. The beauty of this approach is that with enough information, the posterior can still move away from the empirical-Bayes estimated prior.

The dispersion likelihood in the maximum-likelihood estimation is

$$L_g(\phi) = p(y_g\mid\phi)$$

The edgeR weighted shared likelihood is not itself a Bayesian prior, but it can be used to define one. Conceptually:

- The **maximum** of this weighted smoothed likelihood defines the **mean** of the prior. 

- The prior **width** is obtained from the local **curvature** of the weighted smoothed likelihood.

# From frequentist empirical Bayes smoothed shrunk likelihood to Bayesian dispersion prior

Overall, our goal is to match the curvature of the smoothed shrunk (log)-likelihood (around its mode, because it is well behaved) to the curvature of the Bayesian prior, which is a Normal distribution (well behaved everywhere).

- The curvature of the smoothed shrunk (log)-likelihood will be approximated with a quadratic, because it is well behaved around its mode.

- The curvature of a normal distribution is 1/variance by definition.

The following steps will go through this process in detail.

## Smoothed shrunk (log)-likelihood

We take `edgeR` as an example.

`edgeR` estimates an abundance-smoothed shared log-likelihood and weights it through empirical Bayes. For gene $g$, 

$$
w\,\ell_{\mathrm{smooth},g}(\phi),
$$

where

$$
w=\texttt{prior.n}.
$$

In `brmDE`, the dispersion parameter is handled on the log scale. We therefore define the prior kernel directly as a function of $\log(\phi)$:

$$
\log p_g\!\left(\log(\phi)\right)
\propto
w\,\ell_{\mathrm{smooth},g}\left(e^{\log(\phi)}\right).
$$

The maximum of this prior kernel occurs at

$$
\log\left(\phi_{\mathrm{trend},g}\right),
$$

where $\phi_{\mathrm{trend},g}$ is returned by `edgeR` as `trended.dispersion`.

We use the local shape of the log likelihood around its mode, to define the SD of the `brmDE` prior of log phi.

**Intuitive route from weighted smoothed likelihood to prior SD:**

1. Treat the weighted smoothed curve as an unnormalised log-density on the $\log(\phi)$ scale.
2. Exponentiate it to obtain an unnormalised probability density.
3. Normalise it so that its total area is 1.
4. Calculate the mean and SD of that probability distribution.
5. Equivalently, one could sample from the normalised density and calculate the empirical SD.
6. A faster local approximation is to estimate the SD from the curvature around the mode.

## Why curvature can be calculated directly from the weighted log-likelihood

We are calculating the curvature of the **log of the weighted normalised probability density**, as a proxy for the curvature of the prior density itself, prior in this case being likelihood * weight.

Define the weighted smoothed log-likelihood on the log-dispersion scale as

$$
q_g\!\left(\log(\phi)\right)
=
w\,\ell_{\mathrm{smooth},g}\left(e^{\log(\phi)}\right).
$$

The corresponding normalised prior density can be written as

$$
p_g\!\left(\log(\phi)\right)
=
\frac{
\exp\left[
q_g\!\left(\log(\phi)\right)
\right]
}{
Z
},
$$

where $Z$ is the normalising constant.

Taking the logarithm gives

$$
\log p_g\!\left(\log(\phi)\right)
=
q_g\!\left(\log(\phi)\right)-\log Z.
$$

Since $\log Z$ is constant with respect to $\log(\phi)$,

$$
\frac{d^2}{d[\log(\phi)]^2}
\log p_g\!\left(\log(\phi)\right)
=
\frac{d^2}{d[\log(\phi)]^2}
q_g\!\left(\log(\phi)\right).
$$

## The curvature of the `edgeR` likelihood

The smoothed shared log-likelihood is available numerically over a dispersion candidate range, evaluated at chosen candidate values within that range, so we estimate its local curvature from those candidate values.

Suppose there are $G$ genes and $K$ candidate dispersion values.

- `shared.loglik` is a $G \times K$ matrix. Rows correspond to genes and columns correspond to candidate dispersion values.
- `shared.loglik[g, ]` is therefore a numeric vector of length $K$. It gives the y-values of the abundance-smoothed shared log-likelihood for gene $g$ across candidate dispersions.
- `spline.pts` is a numeric vector of length $K$. It gives the candidate-value coordinates corresponding to the $K$ columns of `shared.loglik`.
- `prior.n` is the weight applied to the shared likelihood. In the standard non-robust case it is a scalar; with robust estimation it may be gene-specific.

The x-axis used for the local fit is

$$
x_k
=
\log(\phi_k).
$$

The corresponding y-axis is

$$
y_k
=
w_g
\times
\texttt{shared.loglik}[g,k],
$$

where $w_g$ is `prior.n` for gene $g$ when gene-specific weights are used, or the common scalar `prior.n` otherwise.

Thus, for gene $g$, the local fit uses the paired points

$$
\left(
x_k,
y_k
\right)
=
\left(
\log(\phi_k),
\;
w_g
\times
\texttt{shared.loglik}[g,k]
\right).
$$

## Local quadratic fit

Rather than trying to derive a probability distribution from the smoothed shrunk likelihood and sample from it to get the standard deviation, we use an approximate method.

We estimate the curvature around the expected mode, which tends to be quadratic in shape.

$$
x_{\mathrm{trend},g}
=
\log\left(\phi_{\mathrm{trend},g}\right).
$$

We select a small number of candidate values around the candidate closest to $x_{\mathrm{trend},g}$ and define

$$
z_k
=
x_k
-
x_{\mathrm{trend},g}.
$$

Since the smooth likelihood around its maximum tends to be quadratic in shape, we then fit the local quadratic model

$$
y_k
=
a
+
b z_k
+
c z_k^2.
$$

This can be fitted directly with base R `stats::lm()` because the model is linear in the coefficients $a$, $b$, and $c$.

In R, the design formula is

```r
y[idx] ~ z + I(z^2)
```

where

```r
z <- x[idx] - log(phi_trend_g)
```

and `idx` identifies a small neighbourhood of candidate values around the trended dispersion.

The fitted design matrix has three columns:

$$
\begin{array}{c|ccc}
 & \text{intercept} & z & z^2\\
\hline
1 & 1 & z_1 & z_1^2\\
2 & 1 & z_2 & z_2^2\\
\vdots & \vdots & \vdots & \vdots\\
n & 1 & z_n & z_n^2
\end{array}
$$

and `lm()` estimates

$$
a,\;b,\;c.
$$

The fitted function is therefore a parabola.

The second derivative of the fitted quadratic is

$$
\frac{d^2}{dz^2}
\left(
a+bz+cz^2
\right)
=
2c.
$$

At a valid local maximum,

$$
c<0.
$$

## From a parabola to the curvature

The local curvature of a parabola is defined as:

$$
I_g
=
-2c.
$$

A sharper peak gives larger positive curvature,

$$
\text{sharp peak}
\quad\Longrightarrow\quad
I_g\text{ large},
$$

whereas a flatter peak gives smaller curvature,

$$
\text{flat peak}
\quad\Longrightarrow\quad
I_g\text{ small}.
$$

## Convert curvature into a `brms` prior standard deviation

Matching the curvature of the parabola to the curvature of a Normal distribution, we can derive the SD of the Normal distribution.

For a Normal distribution,

$$
\log p(x)
=
C
-
\frac{
(x-\mu)^2
}{
2\sigma^2
},
$$

so its curvature at the mode is

$$
\frac{1}{\sigma^2}.
$$

Matching the Normal curvature to the local quadratic fit gives

$$
I_g
=
\frac{1}{\sigma_g^2}.
$$

Since

$$
I_g=-2c,
$$

the prior variance is

$$
\sigma_g^2
=
-\frac{1}{2c}.
$$

Therefore the local Normal approximation for the `brmDE` log-dispersion prior is

$$
\log(\phi_g)
\approx
N\left(
\log(\phi_{\mathrm{trend},g}),
-\frac{1}{2c_g}
\right)
$$

with

$$
c_g<0.
$$

The centre of the prior therefore comes from `trended.dispersion`, while its width is derived from the local curvature of the weighted shared edgeR likelihood.

`estimate_dispersion_prior()` writes two columns: `dispersion_prior_log_mean` (`trended.dispersion`) and `dispersion_prior_log_sd` (a width on \(\log\phi\) from the local quadratic of the weighted shared log-likelihood).

# Appendix 1: edgeR calculation of the smoothed weighted likelihood

<div style="background-color: #fff8e1; border-left: 4px solid #e6b800; padding: 0.8em 1em; margin: 1.2em 0; border-radius: 3px;">
<strong>Disclaimer:</strong> the following section should not be taken as a substitute for the official <code>edgeR</code> literature (see references). It is just a convenient complement. For any divergence, the official literature should be considered instead.
</div>

## 1. Each gene has a likelihood curve over candidate dispersions

For gene $g$, with observed counts $y_g$, edgeR evaluates the negative-binomial likelihood

$$
L_g(\phi)
=
p(y_g\mid\phi).
$$

Usually we work with the log-likelihood

$$
\ell_g(\phi)
=
\log p(y_g\mid\phi).
$$

The data $y_g$ are fixed, while $\phi$ varies over candidate dispersion values within a chosen range,

$$
\phi_1,\phi_2,\ldots,\phi_K.
$$

Across genes, this produces a matrix such as

$$
\begin{array}{c|cccc}
 & \phi_1 & \phi_2 & \phi_3 & \phi_4\\
\hline
g_1 & \ell_1(\phi_1) & \ell_1(\phi_2) & \ell_1(\phi_3) & \ell_1(\phi_4)\\
g_2 & \ell_2(\phi_1) & \ell_2(\phi_2) & \ell_2(\phi_3) & \ell_2(\phi_4)\\
g_3 & \ell_3(\phi_1) & \ell_3(\phi_2) & \ell_3(\phi_3) & \ell_3(\phi_4)
\end{array}
$$

Each row is one gene-specific log-likelihood curve over $\phi$.

## 2. edgeR smooths each dispersion column over abundance

Each gene has an abundance value $A_g$.

For one fixed candidate dispersion $\phi_k$, edgeR has the log-likelihood values

$$
\ell_1(\phi_k),
\ell_2(\phi_k),
\ldots,
\ell_G(\phi_k).
$$

It smooths these values against abundance:

$$
A_g
\longrightarrow
\ell_g(\phi_k).
$$

With the default `edgeR` implementation, this smoothing uses local regression through `locfit`.

For every fixed candidate dispersion $\phi_k$, edgeR therefore obtains a smooth function of abundance, $f_k(A)$.

## 3. For one abundance, reconstruct the smoothed log-likelihood curve over dispersion

Consider a particular gene $g$ with abundance $A_g$.

For every candidate dispersion,

$$
\phi_1,\phi_2,\ldots,\phi_K,
$$

edgeR evaluates the corresponding abundance-smoothed function at $A_g$:

$$
f_1(A_g),
f_2(A_g),
\ldots,
f_K(A_g).
$$

These values are placed side by side across dispersion:

$$
\begin{array}{c|cccc}
\phi &
\phi_1 &
\phi_2 &
\cdots &
\phi_K\\
\hline
\text{smoothed log-likelihood} &
\ell_{\mathrm{smooth},g}(\phi_1) &
\ell_{\mathrm{smooth},g}(\phi_2) &
\cdots &
\ell_{\mathrm{smooth},g}(\phi_K)
\end{array}
$$

Together, these values define the function

$$
\ell_{\mathrm{smooth},g}(\phi).
$$

This corresponds to edgeR's internal notation $m_{0g}(\phi)$; for clarity, we use $\ell_{\mathrm{smooth},g}(\phi)$ throughout.

In the edgeR implementation:

- `trended.dispersion` is returned by `estimateDisp()` and gives the abundance-dependent dispersion trend;
- `m0` is the internal matrix of local shared log-likelihoods;
- `prior.n` is the weight applied to the shared likelihood contribution;
- `tagwise.dispersion` is the final gene-specific moderated dispersion estimate.

## 4. The maximum of this curve gives the trended dispersion

edgeR finds the dispersion value at which the smoothed log-likelihood is maximal:

$$
\phi_{\mathrm{trend},g}
=
\arg\max_{\phi}
\ell_{\mathrm{smooth},g}(\phi).
$$

Thus $\ell_{\mathrm{smooth},g}(\phi)$ is the whole curve, whereas $\phi_{\mathrm{trend},g}$ is one value of $\phi$: the dispersion preferred by the across-gene information for a gene with abundance $A_g$.

Repeating this operation over abundance produces the dispersion-abundance trend,

$$
A
\longrightarrow
\phi_{\mathrm{trend}}(A).
$$

## 5. edgeR weights the across-gene log-likelihood curve

We explain the derivation of `prior.df` in a generative fashion. 

What is fixed and already calculated: 

- $d_g$ comes from the design
- $\phi_{\mathrm{trend},g}$ from the abundance smooth. 

What are the master unknown parameters from where everything else is generated:

- The location of the residual deviance $s_0^2$ 
- The tightness around the location, depending on degrees of freedom $d_0$. Large $d_0$ leads to high tightness

**Step 1. Draw the gene's true residual variance.** Not all genes will have the same dispersion deviance (from the common trend $\phi_{\mathrm{trend},g}$), for biological reasons. So each one is given its own true residual variance $\sigma_g^2$, drawn from a scaled inverse chi-square centred on $s_0^2$:

$$
\sigma_g^2
\;\sim\;
\frac{
d_0\,s_0^2
}{
\chi^2_{d_0}
}.
$$



**Step 2. Draw the observed deviance ratio around it.** Given that latent truth, the actual dispersion deviance we observe $s_g^2$ is drawn from a distribution, defined by $\sigma_g^2$.

$$
s_g^2\mid\sigma_g^2
\;\sim\;
\frac{
\sigma_g^2\,\chi^2_{d_g}
}{
d_g
}.
$$

**Eliminate $\sigma_g^2$.** We substitute step 1 into step 2 and the latent $\sigma_g^2$ is integrated out:

$$
s_g^2
\;=\;
\frac{
d_0\,s_0^2
}{
\chi^2_{d_0}
}
\cdot
\frac{
\chi^2_{d_g}
}{
d_g
}
\;=\;
\frac{
d_0\,s_0^2
}{
d_g
}
\cdot
\frac{
\chi^2_{d_g}
}{
\chi^2_{d_0}
}.
$$

The two chi-squares are independent, because step 1 draws $\sigma_g^2$ before the counts of step 2 are realised. A ratio of independent chi-squares, each over its own degrees of freedom, is exactly an $F$:


$$
s_g^2
\;\sim\;
s_0^2\,F_{d_g,\,d_0},
$$

with $s_0^2$ as pure scale and the two degrees of freedom as shape. So the observed vector is marginally a scaled $F$ sample, and no gene's $\sigma_g^2$ is ever computed. Fitting that $F$ is the whole job, and it is `fitFDist()` inside `squeezeVar()`.

### How $s_0^2$ and $d_0$ are recovered

$s_0^2$ enters only as a multiplicative scale, so on the log scale it is a shift. $d_0$ enters through the spread of the $F$. And $d_g$ is already known. They are not solved jointly, but sequentially.

Taking logs and using $\mathbb{E}\left[\log\chi^2_d\right]=\psi(d/2)+\log 2$ and $\mathrm{Var}\left[\log\chi^2_d\right]=\psi'(d/2)$ gives one equation for each:

$$
\mathrm{Var}\left[\log s_g^2\right]
=
\underbrace{\psi'(d_g/2)}_{\text{sampling noise}}
+
\underbrace{\psi'(d_0/2)}_{\text{between-gene spread}},
$$

$$
\mathbb{E}\left[\log s_g^2\right]
=
\log s_0^2
+
\left[\psi(d_g/2)-\log(d_g/2)\right]
-
\left[\psi(d_0/2)-\log(d_0/2)\right].
$$

The variance equation contains $d_0$ alone, because a scale factor cannot change a variance of logs, so $s_0^2$ has dropped out of it. That is what lets $d_0$ be recovered first: subtract the known sampling term and invert the trigamma,

$$
\psi'(d_0/2)
=
\widehat{\mathrm{Var}}\left[\log s_g^2\right]
-
\psi'(d_g/2),
\qquad
d_0
=
2\,\psi'^{-1}\!\left(\cdot\right),
$$

which limma evaluates with `trigammaInverse()`. 

Then, the mean equation gives $\log s_0^2$, since it needs $d_0$ through its digamma term.

This technique is called moment matching, which is a substitute for maximum likelihood.

### From $d_0$ to the weight $w$

edgeR reports $d_0$ as `prior.df`. Dividing one by the other expresses the shared across-gene information,

$$
w
=
\frac{
\texttt{prior.df}
}{
\texttt{nlibs}-\texttt{ncoefs}
}.
$$

So `prior.n` $=2$ means the across-gene trend carries twice as much information as the gene's own counts, and `prior.n` $=0.5$ means half as much.

# Appendix 2: Where edgeR's dispersion and the brms shape differ

**The design.** `estimateDisp()` fits a fixed-effects GLM, so a random-effect term has to be handled somehow, and the two available choices bias $\phi_g$ in opposite directions. *Stripping* it, for example passing `~ dex` while brms fits `~ dex + (1 | cell)`, brings between-cell variance to $\phi_g$, inflating it. Passing the *fixed-effect analogue* (`~ dex + cell`) estimates those cell effects without shrinkage, so they absorb more variance than the mixed model's partially pooled ones will, leaving $\phi_g$ below the conditional shape brms is going to estimate. The target sits between the two:

$$
\phi^{(\sim\,\mathrm{dex}\,+\,\mathrm{cell})}_{\mathrm{edgeR}}
\;\le\;
\phi^{\mathrm{conditional}}_{\mathrm{brms}}
\;\le\;
\phi^{(\sim\,\mathrm{dex})}_{\mathrm{edgeR}},
$$

with the position depending on how hard the mixed model shrinks, which is not known before fitting.

The fixed-effect analogue also costs residual degrees of freedom. On airway's 8 samples, `~ dex + cell` spans 5 columns and leaves $d=3$, whereas `~ dex` leaves $d=6$. This reduction in residual degrees of freedom increases the sampling uncertainty of the gene-wise residual dispersion estimates and can therefore affect the empirically estimated $d_0$. Because $d_0$ sets `prior.n`, and `prior.n` weights the shared curve whose curvature becomes the prior width, the choice of fixed-effect approximation can indirectly change that width.

**The family.** The default is zero-inflated negative binomial. `shape` there is the same parameter as in `negbinomial()`, with the same NB kernel and the same log link; what differs is the estimand. It describes the count component, whereas edgeR fits a plain negative binomial to the marginal counts.

Zero inflation contributes overdispersion quadratic in $\mu$ exactly as the NB dispersion does, so a marginal fit sees the two combined. Matching $\mathrm{Var}(Y) = \mathrm{E}[Y] + \phi\,\mathrm{E}[Y]^2$ against the zero-inflated moments gives

$$
\phi_{\mathrm{marginal}}
=
\frac{\phi+\pi}{1-\pi},
\qquad
\pi=\texttt{zi},
$$

which pins both the direction and the size of that mismatch. edgeR's estimate exceeds the ZINB component's $\phi$ whenever $\pi>0$, by roughly $\pi$ in absolute terms and $\pi/\phi$ in relative terms, so genes with small dispersion are the most distorted. The same expression is why `zi` and `shape` are only weakly separated: the mean-variance relationship constrains the combination, not the pair, leaving the shape of the zero mass as the main information that distinguishes them.

## The Student-t prior confers robstness to misrepresentation of dispersion prior estimates

There are two important situations in which the edgeR dispersion trend can be systematically displaced from the dispersion estimated by `brmDE` (see [Appendix 2](#appendix-2)).

- Because edgeR dispersion estimation cannot fit mixed-effects models, a mixed-effects `brmDE` model must be approximated by a fixed-effects model. For example, `~ treatment + (1 | hospital)` may be approximated by `~ treatment + hospital`. The fixed effects are estimated without partial pooling and can absorb more variation, potentially reducing the residual negative-binomial dispersion.

- A zero-inflated negative-binomial model in `brmDE` must be approximated upstream by an ordinary negative-binomial model. Zero inflation can then be absorbed by the edgeR negative-binomial dispersion, potentially increasing the estimated dispersion.

Student-t has heavier tails, so the gene-wise posterior can leave a misplaced edgeR mode when the likelihood disagrees. The default width still comes from the local curvature of the weighted shared log-likelihood.

# Appendix 3: Examples of how to use R formulas to approximate a mixed-effect model with a fixed-effect one

`edgeR` and `DESeq2` have no random effects. Pass `estimate_dispersion_prior()` the fixed-effect analogue of the mixed model you will fit in `brmDE`: drop the `|` and promote the grouping factor to a fixed term. If the random term also has a slope, include the interaction.

| Mixed-effect model (`brmDE`) | Fixed-effect analogue |
|----|----|
| `~ treatment + (1 \| hospital)` | `~ treatment + hospital` |
| `~ treatment + (1 \| hospital) + (1 \| batch)` | `~ treatment + hospital + batch` |
| `~ treatment + (treatment \| hospital)` | `~ treatment * hospital` |

The last row is the random-slope case: `(treatment | hospital)` is `(1 + treatment | hospital)`, so the analogue needs hospital-specific intercepts and hospital-specific treatment effects, which is `treatment * hospital`.

Do not expand a nested random intercept `(1 | hospital/ward)` as `hospital / ward`. That formula is `hospital + hospital:ward`, which is rank-deficient when every ward belongs to one hospital, and edgeR will not fit it. Use the innermost grouping instead, uniquely coded: `~ treatment + ward`.

# References

Smyth GK (2004). Linear models and empirical Bayes methods for assessing differential expression in microarray experiments. *Statistical Applications in Genetics and Molecular Biology* **3**, 3. https://gksmyth.github.io/pubs/ebayes.pdf. The scaled inverse chi-square hierarchy, the identity $\mathrm{Var}(\log s^2) = \psi'(d/2)$, prior degrees of freedom, and the moderated $t$ / $F$.

Robinson MD, McCarthy DJ, Smyth GK (2010). edgeR: a Bioconductor package for differential expression analysis of digital gene expression data. *Bioinformatics* **26**, 139--140. https://pmc.ncbi.nlm.nih.gov/articles/PMC2796818/. The negative binomial with quadratic overdispersion $\phi_g$.

McCarthy DJ, Chen Y, Smyth GK (2012). Differential expression analysis of multifactor RNA-Seq experiments with respect to biological variation. *Nucleic Acids Research* **40**, 4288--4297. https://pmc.ncbi.nlm.nih.gov/articles/PMC3378882/. Cox--Reid adjusted profile likelihood for $\phi_g$ conditional on a design, and tagwise empirical-Bayes shrinkage towards the trend.

Lund SP, Nettleton D, McCarthy DJ, Smyth GK (2012). Detecting differential expression in RNA-sequence data using quasi-likelihood with shrunken dispersion estimates. *Statistical Applications in Genetics and Molecular Biology* **11**, 8. https://doi.org/10.1515/1544-6115.1826. The QL $F$ test with `df.total = df.residual + df.prior`, applied to a residual dispersion rather than to $\phi_g$.

Phipson B, Lee S, Majewski IJ, Alexander WS, Smyth GK (2016). Robust hyperparameter estimation protects against hypervariable genes and improves power to detect differential expression. *Annals of Applied Statistics* **10**, 946--963. https://doi.org/10.1214/16-AOAS920. `robust = TRUE`, gene-wise `prior.df`, and `prior.df = Inf` under complete shrinkage.

Chen Y, Lun ATL, McCarthy DJ, Ritchie ME, Phipson B, Zhou X, Robinson MD, Smyth GK. `estimateDisp`. In *edgeR User's Guide*, Bioconductor. https://rdrr.io/bioc/edgeR/man/estimateDisp.html. Return value of `tagwise.dispersion`, `prior.df`, and `prior.n`; residual df is not returned.

# Illustration: the local quadratic approximation on airway

The prior width from `estimate_dispersion_prior()` treats edgeR's weighted, abundance-smoothed shared log-likelihood as an unnormalised log-density on $\log(\phi)$, and replaces it with a Normal whose SD matches the local curvature at the mode. That is a *local* quadratic, not a claim that the whole curve is Gaussian.

Below, airway is filtered and the design is the fixed-effect analogue of `~ dex + (1 | cell)`. The black curves are $y = \texttt{prior.n}\times\texttt{shared.loglik}$; the dashed line is $\log(\texttt{trended.dispersion})$; the orange curve is the Laplace Normal $N(\log\phi_{\mathrm{trend}},\sigma^2)$ fitted locally around that mode.

```{r packages, message = FALSE, eval = run_vignette}
# Build the design, drop low-count genes, and run edgeR's own dispersion
# estimation. estimateDisp() supplies prior.n, trended.dispersion, AveLogCPM
# and the plotBCV input used below.
library(edgeR)
library(SummarizedExperiment)
library(ggplot2)

data("airway", package = "airway")

design <- model.matrix(~ dex + cell, data = as.data.frame(colData(airway)))

dge <- DGEList(assay(airway, "counts"))
dge <- dge[filterByExpr(dge, design), , keep.lib.sizes = FALSE]
dge <- normLibSizes(dge)
dge <- estimateDisp(dge, design)
```

```{r shared-loglik, message = FALSE, eval = run_vignette}
# estimateDisp() does not return the shared log-likelihood, so rebuild it:
# Cox--Reid adjusted profile likelihood at each candidate phi, then smooth
# each candidate column against abundance with WLEB.
spline.pts <- seq(-20, 10, length.out = 61)   # wider and finer than edgeR's default
phi <- 0.1 * 2^spline.pts
x <- log(phi)

# Each fit starts from the previous candidate's coefficients; this is what
# estimateDisp() does internally, and the curves are poorly converged without it.
l0 <- matrix(0, nrow(dge), length(phi))
last.beta <- NULL
for (i in seq_along(phi)) {
  out <- adjustedProfileLik(
    phi[[i]],
    y = dge$counts,
    design = design,
    offset = getOffset(dge),
    start = last.beta,
    get.coef = TRUE
  )
  l0[, i] <- out$apl
  last.beta <- out$beta
}

shared.loglik <- WLEB(
  theta = spline.pts,
  loglik = l0,
  covariate = dge$AveLogCPM,
  trend.method = "locfit",
  overall = FALSE,
  individual = FALSE,
  m0.out = TRUE
)$shared.loglik
```

```{r plot-bcv, message = FALSE, eval = run_vignette, fig.width = 6, fig.height = 5}
# Abundance-dispersion trend: points are tagwise BCV = sqrt(phi),
# the line is trended.dispersion, the mode of the curves plotted next.
plotBCV(dge)
```

```{r plot-25, message = FALSE, eval = run_vignette, fig.width = 10, fig.height = 11}
# Take 25 genes evenly spaced in abundance, weight each shared curve by
# prior.n, and fit the local quadratic around log(trended.dispersion) to
# get sigma = sqrt(-1 / (2c)).
idx <- order(dge$AveLogCPM)[round(seq(1, nrow(dge), length.out = 25))]
Y <- rep_len(dge$prior.n, nrow(dge))[idx] * shared.loglik[idx, , drop = FALSE]
x_mode <- log(dge$trended.dispersion[idx])

sigma <- vapply(seq_along(idx), function(i) {
  i0 <- which.min(abs(x - x_mode[[i]]))
  j <- max(1L, i0 - 4L):min(length(x), i0 + 5L)
  z <- x[j] - x_mode[[i]]
  c_hat <- coef(lm(Y[i, j] ~ z + I(z^2)))[[3L]]
  sqrt(-1 / (2 * c_hat))
}, numeric(1))

panel <- sprintf(
  "%s\nlogCPM = %.1f\nσ = %.2f",
  rownames(dge)[idx],
  dge$AveLogCPM[idx],
  sigma
)

curve_df <- data.frame(
  gene = factor(rep(panel, each = length(x)), levels = panel),
  log_phi = rep(x, times = length(idx)),
  y = as.vector(t(Y))
)

# The parabola is drawn on a finer x than the 61 candidates, and clipped to the
# range of the curve so a diverging tail cannot flatten the free y scales.
x_fine <- seq(min(x), max(x), length.out = 201)
normal_df <- do.call(rbind, lapply(seq_along(idx), function(i) {
  y0 <- Y[i, which.min(abs(x - x_mode[[i]]))]
  y_n <- y0 - (x_fine - x_mode[[i]])^2 / (2 * sigma[[i]]^2)
  keep <- y_n >= min(Y[i, ], na.rm = TRUE)
  data.frame(
    gene = factor(panel[[i]], levels = panel),
    log_phi = x_fine[keep],
    y = y_n[keep]
  )
}))

mode_df <- data.frame(gene = factor(panel, levels = panel), log_phi_trend = x_mode)

ggplot(curve_df, aes(log_phi, y)) +
  geom_line(aes(colour = "edgeR weighted shared log-likelihood")) +
  geom_line(data = normal_df, aes(colour = "Laplace Normal"), linetype = 22) +
  geom_vline(
    data = mode_df,
    aes(xintercept = log_phi_trend),
    linetype = 2,
    colour = "grey40"
  ) +
  scale_colour_manual(values = c(
    "edgeR weighted shared log-likelihood" = "black",
    "Laplace Normal" = "#d95f02"
  )) +
  facet_wrap(~ gene, scales = "free_y", ncol = 5) +
  labs(x = expression(log(phi)), y = "prior.n × shared.loglik", colour = NULL) +
  theme_bw() +
  theme(
    aspect.ratio = 1,
    strip.text = element_text(size = 7),
    axis.text = element_text(size = 6),
    legend.position = "bottom"
  )
```
