---
title: "Integrating fast frequentist empirical-Bayes gene-expression dispersion estimation as a prior"
author: "Stefano Mangiola"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating fast frequentist empirical-Bayes gene-expression dispersion estimation as a prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 3.8,
  eval = FALSE
)
run_vignette <-
  requireNamespace("airway", quietly = TRUE) &&
  requireNamespace("edgeR", quietly = TRUE) &&
  requireNamespace("limma", quietly = TRUE) &&
  requireNamespace("tidybulk", quietly = TRUE)
```

`brmDE`, although very flexible and capable of mixed-effect models, non-parametric smooth/spline models, and alternative distributions such as NB and ZINB, fits one gene at a time to prioritise parallelisation. A single-gene likelihood cannot borrow strength across features, which is exactly what makes empirical-Bayes estimation of feature-wise negative-binomial dispersion useful when the number of replicates is small.

Frequentist methods such as `edgeR` estimate gene-wise dispersion through empirical Bayes very efficiently, although generally within more constrained fixed-effect model structures.

The key idea is to estimate gene dispersion cheaply with empirical Bayes, for example through `edgeR` or `DESeq2`, and use the resulting across-gene dispersion information as an approximate prior for `brmDE`. With enough information in the gene's own counts, the posterior can still move away from the upstream empirical-Bayes estimate.

The conceptual transition is from a likelihood over candidate dispersion values to a prior kernel for dispersion. For fixed observed data, the likelihood

$$
L_g(\phi)
=
p(y_g\mid\phi)
$$

is a function of the candidate parameter value $\phi$. It is not itself a probability density for $\phi$. However, the empirical-Bayes machinery in `edgeR` constructs a weighted, abundance-smoothed log-likelihood contribution that represents information borrowed across genes. In `brmDE`, we use this weighted shared log-likelihood as an empirical prior kernel for the parameter that is actually modelled, $\log(\phi)$.

This is an explicit modelling approximation: the edgeR weighted shared likelihood is not itself a Bayesian prior, but it can be used to define one.

The maximum of this weighted smoothed likelihood defines the **mode** of the prior. Under the local Normal approximation developed below, the mode is also used as the prior mean. The prior width is obtained from the local curvature of the same weighted smoothed likelihood.

Because the smoothed dispersion trend is returned directly by `edgeR` as `trended.dispersion`, the less intuitive part is the transition from the weighted shared likelihood to the prior SD. That calculation is described below.

# From smoothed shrunk likelihood to log-scale SD

`edgeR` estimates an abundance-smoothed shared log-likelihood and weights it through empirical Bayes. For gene $g$, we denote this as

$$
w\,\ell_{\mathrm{smooth},g}(\phi),
$$

where

$$
w=\texttt{prior.n}.
$$

In `brmDE`, the dispersion parameter is handled on the log scale. We therefore define the prior kernel directly as a function of $\log(\phi)$:

$$
\log p_g\!\left(\log(\phi)\right)
\propto
w\,\ell_{\mathrm{smooth},g}\left(e^{\log(\phi)}\right).
$$

This avoids introducing an unnecessary change-of-variable step from a density defined with respect to $d\phi$ to one defined with respect to $d\log(\phi)$. We are explicitly defining the prior on the scale used by `brmDE`.

The maximum of this prior kernel occurs at

$$
\log\left(\phi_{\mathrm{trend},g}\right),
$$

where $\phi_{\mathrm{trend},g}$ is returned by `edgeR` as `trended.dispersion`.

The weighted edgeR curve is not itself a Bayesian prior density. For constructing the `brmDE` prior, we use its local shape around the maximum.

**Intuitive route from weighted smoothed likelihood to prior SD:**

1. Treat the weighted smoothed curve as an unnormalised log-density on the $\log(\phi)$ scale.
2. Exponentiate it to obtain an unnormalised probability density.
3. Normalise it so that its total area is 1.
4. Calculate the mean and SD of that probability distribution.
5. Equivalently, one could sample from the normalised density and calculate the empirical SD.
6. A faster local approximation is to estimate the SD from the curvature around the mode.

## Why curvature can be calculated directly from the weighted log-likelihood

We are calculating the curvature of the **log of the normalised probability density**, as a proxy for the curvature of the probability density itself.

Define the weighted smoothed log-likelihood on the log-dispersion scale as

$$
q_g\!\left(\log(\phi)\right)
=
w\,\ell_{\mathrm{smooth},g}\left(e^{\log(\phi)}\right).
$$

The corresponding normalised prior density can be written as

$$
p_g\!\left(\log(\phi)\right)
=
\frac{
\exp\left[
q_g\!\left(\log(\phi)\right)
\right]
}{
Z
},
$$

where $Z$ is the normalising constant.

Taking the logarithm gives

$$
\log p_g\!\left(\log(\phi)\right)
=
q_g\!\left(\log(\phi)\right)-\log Z.
$$

Since $\log Z$ is constant with respect to $\log(\phi)$,

$$
\frac{d^2}{d[\log(\phi)]^2}
\log p_g\!\left(\log(\phi)\right)
=
\frac{d^2}{d[\log(\phi)]^2}
q_g\!\left(\log(\phi)\right).
$$

Therefore,

$$
\text{curvature of the log normalised prior}
=
\text{curvature of the weighted smoothed log-likelihood}.
$$

This means that the conceptual exponentiation and normalisation steps do not need to be carried out explicitly when estimating curvature.

## Numerical curvature from the edgeR likelihood grid

The abundance-smoothed shared log-likelihood is available numerically over a finite dispersion grid, so we estimate its local curvature from those grid values.

Suppose there are $G$ genes and $K$ dispersion-grid points.

- `shared.loglik` is a $G \times K$ matrix. Rows correspond to genes and columns correspond to dispersion-grid points.
- `shared.loglik[g, ]` is therefore a numeric vector of length $K$. It gives the y-values of the abundance-smoothed shared log-likelihood for gene $g$ across candidate dispersions.
- `spline.pts` is a numeric vector of length $K$. It gives the grid coordinates corresponding to the $K$ columns of `shared.loglik`.
- `prior.n` is the weight applied to the shared likelihood. In the standard non-robust case it is a scalar; with robust estimation it may be gene-specific.

The x-axis used for the local fit is

$$
x_k
=
\log(\phi_k).
$$

The corresponding y-axis is

$$
y_k
=
w_g
\times
\texttt{shared.loglik}[g,k],
$$

where $w_g$ is `prior.n` for gene $g` when gene-specific weights are used, or the common scalar `prior.n` otherwise.

Thus, for gene $g$, the local fit uses the paired points

$$
\left(
x_k,
y_k
\right)
=
\left(
\log(\phi_k),
\;
w_g
\times
\texttt{shared.loglik}[g,k]
\right).
$$

## Local quadratic fit

Rather than trying to derive probability distribution from smooth shrinked likelihood and sample from it to get the standard deviation, we use an approximate method. 

We estimate the curvature around the expected mode

$$
x_{\mathrm{trend},g}
=
\log\left(\phi_{\mathrm{trend},g}\right).
$$

We select a small number of grid points around the grid location closest to $x_{\mathrm{trend},g}$ and define

$$
z_k
=
x_k
-
x_{\mathrm{trend},g}.
$$

Since, smooth likelihood around their maximum tend to be quadratic in shape, we then fit the local quadratic model

$$
y_k
=
a
+
b z_k
+
c z_k^2.
$$

This can be fitted directly with base R `stats::lm()` because the model is linear in the coefficients $a$, $b$, and $c$.

In R, the design formula is

```r
y[idx] ~ z + I(z^2)
```

where

```r
z <- x[idx] - log(phi_trend_g)
```

and `idx` identifies a small neighbourhood of grid points around the trended dispersion.

The fitted design matrix has three columns:

$$
\begin{array}{c|ccc}
 & \text{intercept} & z & z^2\\
\hline
1 & 1 & z_1 & z_1^2\\
2 & 1 & z_2 & z_2^2\\
\vdots & \vdots & \vdots & \vdots\\
n & 1 & z_n & z_n^2
\end{array}
$$

and `lm()` estimates

$$
a,\;b,\;c.
$$

The fitted function is therefore a parabola, not a straight line.

The second derivative of the fitted quadratic is

$$
\frac{d^2}{dz^2}
\left(
a+bz+cz^2
\right)
=
2c.
$$

At a valid local maximum,

$$
c<0.
$$

We therefore define the positive local curvature as

$$
I_g
=
-2c.
$$

This estimates

$$
I_g
=
-
\left.
\frac{d^2}{d[\log(\phi)]^2}
\left[
w\,\ell_{\mathrm{smooth},g}(\phi)
\right]
\right|_{\phi=\phi_{\mathrm{trend},g}}.
$$

A sharper peak gives larger positive curvature,

$$
\text{sharp peak}
\quad\Longrightarrow\quad
I_g\text{ large},
$$

whereas a flatter peak gives smaller curvature,

$$
\text{flat peak}
\quad\Longrightarrow\quad
I_g\text{ small}.
$$

## Convert curvature into a `brms` prior standard deviation

The final step is a local Normal, or Laplace, approximation.

Near its mode, a sufficiently smooth log-density can be approximated by a quadratic. For a Normal distribution,

$$
\log p(x)
=
C
-
\frac{
(x-\mu)^2
}{
2\sigma^2
},
$$

so its curvature at the mode is

$$
\frac{1}{\sigma^2}.
$$

Matching the Normal curvature to the local quadratic fit gives

$$
I_g
=
\frac{1}{\sigma_g^2}.
$$

Since

$$
I_g=-2c,
$$

the prior variance is

$$
\sigma_g^2
=
-\frac{1}{2c}.
$$

Therefore the local Normal approximation for the `brmDE` log-dispersion prior is

$$
\boxed{
\log(\phi_g)
\approx
N\left(
\log(\phi_{\mathrm{trend},g}),
-\frac{1}{2c_g}
\right)
}
$$

with

$$
c_g<0.
$$

The centre of the prior therefore comes from `trended.dispersion`, while its width is derived from the local curvature of the weighted shared edgeR likelihood.

In `brmDE` the default width is the Laplace SD. `estimate_dispersion_prior()` writes two columns: `dispersion_prior_log_mean` (`trended.dispersion`) and `dispersion_prior_log_sd` (a width on \(\log\phi\)). `method = "curvature"` (default) fills that SD from the local quadratic of the weighted shared log-likelihood; `method = "degrees_freedom"` fills it from the trigamma SD of the moderated df. Pass both names to `estimate_gene()` / `estimate()`. The location is the offset on `shape ~ 1 + offset(log(1/dispersion_prior_log_mean))`. The intercept prior is Student-t, with scale chosen so that its standard deviation equals that log-scale \(\sigma_g\).

# Appendix 1: edgeR calculation of the smoothed weighted likelihood

**
disclamer: the following section should not be taken as substitute of the official `edgeR` literature (see references). It is just a convenient complement. For any divergence, the official literature should be considered instead.**

## 1. Each gene has a likelihood curve over candidate dispersions

For gene $g$, with observed counts $y_g$, edgeR evaluates the negative-binomial likelihood

$$
L_g(\phi)
=
p(y_g\mid\phi).
$$

Usually we work with the log-likelihood

$$
\ell_g(\phi)
=
\log p(y_g\mid\phi).
$$

The data $y_g$ are fixed, while $\phi$ varies over a grid of candidate dispersion values,

$$
\phi_1,\phi_2,\ldots,\phi_K.
$$

Across genes, this produces a matrix such as

$$
\begin{array}{c|cccc}
 & \phi_1 & \phi_2 & \phi_3 & \phi_4\\
\hline
g_1 & \ell_1(\phi_1) & \ell_1(\phi_2) & \ell_1(\phi_3) & \ell_1(\phi_4)\\
g_2 & \ell_2(\phi_1) & \ell_2(\phi_2) & \ell_2(\phi_3) & \ell_2(\phi_4)\\
g_3 & \ell_3(\phi_1) & \ell_3(\phi_2) & \ell_3(\phi_3) & \ell_3(\phi_4)
\end{array}
$$

Each row is one gene-specific log-likelihood curve over $\phi$.

## 2. edgeR smooths each dispersion column over abundance

Each gene also has an abundance value

$$
A_g.
$$

For one fixed candidate dispersion $\phi_k$, edgeR has the values

$$
\ell_1(\phi_k),
\ell_2(\phi_k),
\ldots,
\ell_G(\phi_k).
$$

It smooths these values against abundance:

$$
A_g
\longrightarrow
\ell_g(\phi_k).
$$

With the default edgeR implementation, this smoothing uses local regression through `locfit`.

For every fixed candidate dispersion $\phi_k$, edgeR therefore obtains a smooth function of abundance,

$$
f_k(A).
$$

For example,

$$
f_1(A)
$$

describes the expected log-likelihood at dispersion $\phi_1$ for a gene of abundance $A$, while

$$
f_2(A)
$$

does the same for $\phi_2$, and so on.

## 3. For one abundance, reconstruct the smoothed log-likelihood curve over dispersion

Consider a particular gene $g$ with abundance

$$
A_g.
$$

For every candidate dispersion,

$$
\phi_1,\phi_2,\ldots,\phi_K,
$$

edgeR evaluates the corresponding abundance-smoothed function at $A_g$:

$$
f_1(A_g),
f_2(A_g),
\ldots,
f_K(A_g).
$$

These values are placed side by side across dispersion:

$$
\begin{array}{c|cccc}
\phi &
\phi_1 &
\phi_2 &
\cdots &
\phi_K\\
\hline
\text{smoothed log-likelihood} &
\ell_{\mathrm{smooth},g}(\phi_1) &
\ell_{\mathrm{smooth},g}(\phi_2) &
\cdots &
\ell_{\mathrm{smooth},g}(\phi_K)
\end{array}
$$

Together, these values define the function

$$
\ell_{\mathrm{smooth},g}(\phi).
$$

This corresponds to edgeR's internal notation $m_{0g}(\phi)$; for clarity, we use $\ell_{\mathrm{smooth},g}(\phi)$ throughout.

In the edgeR implementation:

- `trended.dispersion` is returned by `estimateDisp()` and gives the abundance-dependent dispersion trend;
- `m0` is the internal matrix of local shared log-likelihoods;
- `prior.n` is the weight applied to the shared likelihood contribution;
- `tagwise.dispersion` is the final gene-specific moderated dispersion estimate.

Thus,

$$
\ell_{\mathrm{smooth},g}(\phi)
\longleftrightarrow
\texttt{m0},
$$

and

$$
w
\longleftrightarrow
\texttt{prior.n}.
$$

The abundance-dependent maximum of the shared likelihood corresponds to

$$
\phi_{\mathrm{trend},g}
\longleftrightarrow
\texttt{trended.dispersion}.
$$

`m0` is an internal intermediate quantity and is not normally retained as a component of the `DGEList` returned by `estimateDisp()`. If `brmDE` needs the full shared likelihood curve, the relevant edgeR internal calculation must therefore be called, reproduced, or wrapped so that this matrix is retained.

The whole curve

$$
\ell_{\mathrm{smooth},g}(\phi)
$$

can be interpreted as:

> According to the likelihood behaviour of genes with similar abundance, how compatible is each candidate dispersion $\phi$?

## 4. The maximum of this curve gives the trended dispersion

edgeR finds the dispersion value at which the smoothed log-likelihood is maximal:

$$
\phi_{\mathrm{trend},g}
=
\arg\max_{\phi}
\ell_{\mathrm{smooth},g}(\phi).
$$

Thus $\ell_{\mathrm{smooth},g}(\phi)$ is the whole curve, whereas $\phi_{\mathrm{trend},g}$ is one value of $\phi$: the dispersion preferred by the across-gene information for a gene with abundance $A_g$.

Repeating this operation over abundance produces the dispersion-abundance trend,

$$
A
\longrightarrow
\phi_{\mathrm{trend}}(A).
$$

## 5. edgeR weights the across-gene log-likelihood curve

edgeR estimates prior degrees of freedom through its empirical-Bayes procedure and converts these into the likelihood weight used by the weighted-likelihood machinery.

In the standard non-robust implementation,

$$
\texttt{prior.n}
=
\frac{
\texttt{prior.df}
}{
\texttt{nlibs}-\texttt{ncoefs}
}.
$$

With robust estimation, `prior.df` and therefore `prior.n` may be gene-specific.

The weight controls how strongly the across-gene information contributes relative to the likelihood of the individual gene.

A larger weight means stronger shrinkage towards the across-gene dispersion trend.

The weighted curve is

$$
w\,\ell_{\mathrm{smooth},g}(\phi).
$$

Multiplying the log-likelihood by a positive weight does not change the location of its maximum:

$$
\arg\max_{\phi}
w\,\ell_{\mathrm{smooth},g}(\phi)
=
\phi_{\mathrm{trend},g}.
$$

It does change the sharpness of the curve.

A larger $w$ amplifies differences between candidate dispersion values and produces a sharper curve. A smaller $w$ produces a flatter curve.

On the ordinary likelihood scale, multiplying the log-likelihood by $w$ is equivalent to raising the likelihood to the power $w$:

$$
\exp\left[
w\,\ell_{\mathrm{smooth},g}(\phi)
\right]
=
\left[
\exp\left(
\ell_{\mathrm{smooth},g}(\phi)
\right)
\right]^w.
$$

For the edgeR tagwise dispersion, this weighted across-gene component is combined with the gene's own log-likelihood.

For `brmDE`, the gene's own likelihood should not be included in the prior because the same gene will subsequently be fitted by `brms`. The relevant prior information is therefore the weighted across-gene curve

$$
w\,\ell_{\mathrm{smooth},g}(\phi).
$$

# Appendix 2: Where edgeR's dispersion and the brms shape differ

**The design.** `estimateDisp()` fits a fixed-effects GLM, so a random-effect term has to be handled somehow, and the two available choices bias $\phi_g$ in opposite directions. *Stripping* it, for example passing `~ dex` while brms fits `~ dex + (1 | cell)`, leaves the between-cell variance nowhere to go but $\phi_g$, inflating it. Passing the *fixed-effect analogue* (`~ dex + cell`) estimates those cell effects without shrinkage, so they absorb more variance than the mixed model's partially pooled ones will, leaving $\phi_g$ below the conditional shape brms is going to estimate. The target sits between the two:

$$
\phi^{(\sim\,\mathrm{dex}\,+\,\mathrm{cell})}_{\mathrm{edgeR}}
\;\le\;
\phi^{\mathrm{conditional}}_{\mathrm{brms}}
\;\le\;
\phi^{(\sim\,\mathrm{dex})}_{\mathrm{edgeR}},
$$

with the position depending on how hard the mixed model shrinks, which is not known before fitting.

The fixed-effect analogue also costs residual degrees of freedom. On airway's 8 samples, `~ dex + cell` spans 5 columns and leaves $d=3$, whereas `~ dex` leaves $d=6$. This reduction in residual degrees of freedom increases the sampling uncertainty of the gene-wise residual dispersion estimates and can therefore affect the empirically estimated $d_0$. Because the `brmDE` prior width is ultimately calibrated from $d_0$, the choice of fixed-effect approximation can indirectly change the prior width.

**The family.** The default is zero-inflated negative binomial. `shape` there is the same parameter as in `negbinomial()`, with the same NB kernel and the same log link; what differs is the estimand. It describes the count component, whereas edgeR fits a plain negative binomial to the marginal counts.

Zero inflation contributes overdispersion quadratic in $\mu$ exactly as the NB dispersion does, so a marginal fit sees the two combined. Matching $\mathrm{Var}(Y) = \mathrm{E}[Y] + \phi\,\mathrm{E}[Y]^2$ against the zero-inflated moments gives

$$
\phi_{\mathrm{marginal}}
=
\frac{\phi+\pi}{1-\pi},
\qquad
\pi=\texttt{zi},
$$

which pins both the direction and the size of that mismatch. edgeR's estimate exceeds the ZINB component's $\phi$ whenever $\pi>0$, by roughly $\pi$ in absolute terms and $\pi/\phi$ in relative terms, so genes with small dispersion are the most distorted. The same expression is why `zi` and `shape` are only weakly separated: the mean-variance relationship constrains the combination, not the pair, leaving the shape of the zero mass as the main information that distinguishes them.

## The Student-t prior

There are two important situations in which the edgeR dispersion trend can be systematically displaced from the dispersion estimated by `brmDE` (see [Appendix 1](#appendix-1)).

- Because edgeR dispersion estimation cannot fit mixed-effects models, a mixed-effects `brmDE` model must be approximated by a fixed-effects model. For example, `~ treatment + (1 | hospital)` may be approximated by `~ treatment + hospital`. The fixed effects are estimated without partial pooling and can absorb more variation, potentially reducing the residual negative-binomial dispersion.

- A zero-inflated negative-binomial model in `brmDE` must be approximated upstream by an ordinary negative-binomial model. Zero inflation can then be absorbed by the edgeR negative-binomial dispersion, potentially increasing the estimated dispersion.

For these cases, call `estimate_dispersion_prior(..., method = "degrees_freedom")`: the same Student-t intercept prior, but with `dispersion_prior_log_sd` filled from the log-scale SD \(\sqrt{\psi'(d_{\mathrm{eff}}/2)}\) implied by the **moderated** effective df \(d_{\mathrm{eff}} = d_{\mathrm{residual}} + d_0\) (`prior.df` from `squeezeVar()` / `estimateDisp()`). tidybulk writes the df as `dispersion_degrees_freedom`; convert it with the same trigamma formula, or let `estimate_dispersion_prior()` do it. The default `method = "curvature"` is the local Laplace approximation of the weighted smoothed shared log-likelihood derived above.


# Appendix 3: Examples of how to use R formulas approximate a mixed-effect model with a fixed-effect one

`edgeR` and `DESeq2` have no random effects. To obtain the most relevant dispersion estimate, we can approximate the model fitted in `brmDE` using its fixed-effect analogue: drop the `|` and promote the grouping factor to a fixed term. If the random term also has a slope, include the interaction.

| Mixed-effect model (`brmDE`) | Fixed-effect analogue (`estimate_dispersion()`) |
|----|----|
| `~ treatment + (1 \| hospital)` | `~ treatment + hospital` |
| `~ treatment + (1 \| hospital) + (1 \| batch)` | `~ treatment + hospital + batch` |
| `~ treatment + (treatment \| hospital)` | `~ treatment * hospital` |

The last row is the random-slope case: `(treatment | hospital)` is `(1 + treatment | hospital)`, so the analogue needs hospital-specific intercepts and hospital-specific treatment effects, which is `treatment * hospital`.

Do not expand a nested random intercept `(1 | hospital/ward)` as `hospital / ward`. That formula is `hospital + hospital:ward`, which is rank-deficient when every ward belongs to one hospital, and edgeR will not fit it. Use the innermost grouping instead, uniquely coded: `~ treatment + ward`.

# References

Smyth GK (2004). Linear models and empirical Bayes methods for assessing differential expression in microarray experiments. *Statistical Applications in Genetics and Molecular Biology* **3**, 3. https://gksmyth.github.io/pubs/ebayes.pdf. The scaled inverse chi-square hierarchy, the identity $\mathrm{Var}(\log s^2) = \psi'(d/2)$, prior degrees of freedom, and the moderated $t$ / $F$.

Robinson MD, McCarthy DJ, Smyth GK (2010). edgeR: a Bioconductor package for differential expression analysis of digital gene expression data. *Bioinformatics* **26**, 139--140. https://pmc.ncbi.nlm.nih.gov/articles/PMC2796818/. The negative binomial with quadratic overdispersion $\phi_g$.

McCarthy DJ, Chen Y, Smyth GK (2012). Differential expression analysis of multifactor RNA-Seq experiments with respect to biological variation. *Nucleic Acids Research* **40**, 4288--4297. https://pmc.ncbi.nlm.nih.gov/articles/PMC3378882/. Cox--Reid adjusted profile likelihood for $\phi_g$ conditional on a design, and tagwise empirical-Bayes shrinkage towards the trend.

Lund SP, Nettleton D, McCarthy DJ, Smyth GK (2012). Detecting differential expression in RNA-sequence data using quasi-likelihood with shrunken dispersion estimates. *Statistical Applications in Genetics and Molecular Biology* **11**, 8. https://doi.org/10.1515/1544-6115.1826. The QL $F$ test with `df.total = df.residual + df.prior`, applied to a residual dispersion rather than to $\phi_g$.

Phipson B, Lee S, Majewski IJ, Alexander WS, Smyth GK (2016). Robust hyperparameter estimation protects against hypervariable genes and improves power to detect differential expression. *Annals of Applied Statistics* **10**, 946--963. https://doi.org/10.1214/16-AOAS920. `robust = TRUE`, gene-wise `prior.df`, and `prior.df = Inf` under complete shrinkage.

Chen Y, Lun ATL, McCarthy DJ, Ritchie ME, Phipson B, Zhou X, Robinson MD, Smyth GK. `estimateDisp`. In *edgeR User's Guide*, Bioconductor. https://rdrr.io/bioc/edgeR/man/estimateDisp.html. Return value of `tagwise.dispersion`, `prior.df`, and `prior.n`; residual df is not returned.

# Illustration: the local quadratic approximation on airway

The default prior width (`estimate_dispersion_prior(..., method = "curvature")`) treats edgeR's weighted, abundance-smoothed shared log-likelihood as an unnormalised log-density on $\log(\phi)$, and replaces it with a Normal whose SD matches the local curvature at the mode. That is a *local* quadratic, not a claim that the whole curve is Gaussian.

The following uses the airway dexamethasone experiment to look at that approximation. After dropping low-count genes, the shared log-likelihood is rebuilt on the same Cox--Reid grid and `WLEB()` smooth that `estimateDisp()` uses. `plotBCV()` shows the abundance–dispersion trend those curves sit on. Then 25 genes spanning the abundance range are drawn as

$$
x_k = \log(\phi_k),
\qquad
y_k = \texttt{prior.n}_g \times \texttt{shared.loglik}[g, k],
$$

the same `plot(x, y)` `estimate_dispersion_prior()` builds internally. The dashed vertical line is $\log(\texttt{trended.dispersion})$, the mode at which the quadratic is fitted. The orange curve is the Laplace Normal on that same log-density scale: $N(\log\phi_{\mathrm{trend}}, \sigma^2)$ with $\sigma$ from the local curvature, shifted to agree with $y$ at the mode.

```{r packages, message = FALSE, eval = run_vignette}
library(brmDE)
library(tidybulk)
library(tidySummarizedExperiment)
library(SummarizedExperiment)
library(ggplot2)
library(dplyr)

data("airway", package = "airway")

airway <- airway |>
  mutate(
    dex = relevel(factor(dex), ref = "untrt"),
    cell = droplevels(cell)
  )

se <- airway |>
  keep_abundant(formula_design = ~ dex)

nrow(se)
```

`keep_abundant()` drops the low-count genes (`identify_abundant()` would only flag them). The design is the fixed-effect analogue of `~ dex + (1 | cell)`.

```{r shared-loglik, message = FALSE, eval = run_vignette}
design <- model.matrix(
  ~ dex + cell,
  data = as.data.frame(colData(se))
)
shared <- brmDE:::edger_shared_loglik(
  assay(se, "counts"),
  design
)

# x <- log(0.1 * 2^spline.pts). estimateDisp() uses 21 points on
# spline.pts in [-10, 10]; we use 61 on [-20, 10] so the curve is not
# stuck on two coarse ticks.
x <- log(shared$phi)
```

edgeR's `plotBCV()` is the abundance–dispersion trend on the same filtered matrix and design. BCV is \(\sqrt{\phi}\). Each point is a gene's tagwise estimate; the line is `trended.dispersion`, the mode of the shared curve in the facets below. On airway that trend is real but narrow, which is why those peaks sit near the same \(\log(\phi)\).

```{r plot-bcv, message = FALSE, eval = run_vignette, fig.width = 6, fig.height = 5}
dge <- edgeR::DGEList(SummarizedExperiment::assay(se, "counts"))
dge <- edgeR::calcNormFactors(dge)
dge <- edgeR::estimateDisp(dge, design)
edgeR::plotBCV(dge)
```

Twenty-five genes spanning abundance: order the filtered genes by edgeR `AveLogCPM`, split that list into 25 equal slices, and take the middle gene of each slice, except the last slice, which uses the most abundant gene so the range is not truncated. Facets run from lowest to highest abundance. The strip shows CPM (`2^AveLogCPM`), not the log2 value. The black line is `y = prior.n * shared.loglik[g, ]`; the orange line is the Normal approximation from that gene's local curvature.

```{r plot-25, message = FALSE, eval = run_vignette}
ave_log_cpm <- edgeR::aveLogCPM(assay(se, "counts"))
cpm <- 2^ave_log_cpm
ord <- order(ave_log_cpm)
n <- length(ord)
breaks <- floor(seq(1, n + 1, length.out = 26))
idx <- vapply(seq_len(25), function(i) {
  sl <- ord[seq.int(breaks[[i]], breaks[[i + 1L]] - 1L)]
  if (i == 25L) {
    sl[[length(sl)]]
  } else {
    sl[[ceiling(length(sl) / 2)]]
  }
}, integer(1))

genes <- rownames(se)[idx]
Y <- shared$prior.n[idx] * shared$shared.loglik[idx, , drop = FALSE]
x_mode <- log(shared$trended.dispersion[idx])
sigma <- vapply(seq_along(idx), function(i) {
  brmDE:::dispersion_quadratic_log_sd(x, Y[i, ], x_mode[[i]])
}, numeric(1))
panel <- sprintf(
  "%s\nCPM = %s\nσ = %.2f",
  genes,
  format(round(cpm[idx], 2), trim = TRUE, scientific = FALSE),
  sigma
)

plot_df <- tibble(
  gene = factor(
    rep(panel, each = length(x)),
    levels = panel
  ),
  log_phi = rep(x, times = length(idx)),
  y = as.vector(t(Y))
)

x_fine <- seq(min(x), max(x), length.out = 201)
approx_df <- lapply(seq_along(idx), function(i) {
  y <- Y[i, ]
  mu <- x_mode[[i]]
  s <- sigma[[i]]
  y0 <- y[[which.min(abs(x - mu))]]
  y_n <- y0 - (x_fine - mu)^2 / (2 * s^2)
  keep <- is.finite(y_n) & y_n >= min(y, na.rm = TRUE)
  tibble(
    gene = factor(panel[[i]], levels = panel),
    log_phi = x_fine[keep],
    y = y_n[keep]
  )
}) |>
  bind_rows()

trend_df <- tibble(
  gene = factor(panel, levels = panel),
  log_phi_trend = x_mode
)
```

```{r plot-25-facets, message = FALSE, eval = run_vignette, fig.width = 8, fig.height = 16}

ggplot(plot_df, aes(log_phi, y)) +
  geom_line(aes(colour = "edgeR weighted shared log-likelihood")) +
  geom_line(
    data = approx_df,
    aes(colour = "Laplace Normal"),
    linetype = 22
  ) +
  geom_vline(
    data = trend_df,
    aes(xintercept = log_phi_trend),
    linetype = 2,
    colour = "grey40"
  ) +
  scale_colour_manual(
    values = c(
      "edgeR weighted shared log-likelihood" = "black",
      "Laplace Normal" = "#d95f02"
    )
  ) +
  facet_wrap(~ gene, scales = "free_y", ncol = 5) +
  labs(
    x = expression(log(phi)),
    y = "prior.n × shared.loglik",
    colour = NULL
  ) +
  theme_bw() +
  theme(
    strip.text = element_text(size = 7),
    axis.text = element_text(size = 6),
    legend.position = "bottom"
  )
```
